## Code to extract habitat data from AusTraits

Using a combination of methods

- for large datasets that are entirely rainforest plants, assign all taxa as having habitat "rainforest"
- for other datasets, search location_properties for terms related to habitat

This forms a dataset that has been input into the threatened species dataset, but should be re-run as AusTraits grows

```{r}
rainforest <- (austraits %>%
  extract_dataset(c("ATRP_2020", "Cooper_2013", "Cooper_2004")))$traits %>%
  distinct(taxon_name, dataset_id) %>%
  group_by(taxon_name) %>%
  mutate(source_id = paste0(dataset_id, collapse = "; ")) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "rainforest")

rainforest2 <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("rainforest", "tropical moist forest", "rain forest")))$traits %>%
  filter(!dataset_id %in% c("ATRP_2020", "Cooper_2013", "Cooper_2004")) %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "rainforest")

aquatic <- (austraits %>%
  extract_trait("plant_growth_substrate") %>%
  extract_data(table = "traits", col = "value", col_value = "aquatic"))$traits %>%
  tidyr::separate_longer_delim(cols = value, delim = " ") %>%
  distinct(taxon_name, value, dataset_id) %>%
  filter(str_detect(value, "aquatic")) %>%
  group_by(taxon_name, value) %>%
  mutate(source_id = paste0(dataset_id, collapse = "; ")) %>%
  ungroup() %>%
  distinct(taxon_name, value, source_id) %>%
  group_by(taxon_name, source_id) %>%
  mutate(habitat = paste0(value, collapse = " ")) %>%
  ungroup() %>%
  distinct(taxon_name, habitat, source_id)

wet_sclerophyll <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = "wet sclerophyll"))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "forest_wet_sclerophyll")

dry_sclerophyll <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = "dry sclerophyll"))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "forest_dry_sclerophyll")

wet_heath <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = "wet heath"))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "heath_wet")

heath <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = "heath"))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "heath")

grassland <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("grassland", "spinifex")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "grassland")

savanna <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("savanna")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "grassland")

bogs_fens <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("bog", "fen", "marsh")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "bog")

shrubland <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("shrubland", "scrub", "mallee")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "shrubland")

eucalypt_forest <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("eucalypt forest")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "forest_eucalypt")

closed_forest <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("closed forest")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "forest_closed")

alpine <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("alpine")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "alpine")

open_forest <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("open forest")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "forest_open")

dry_forest <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("dry forest")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "forest_dry")

wet_forest <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("wet forest")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "forest_wet")


woodland <- (austraits %>%
  extract_data(table = "locations", col = "location_property", col_value = c("habitat", "description")) %>%
  extract_data(table = "locations", col = "value", col_value = c("woodland")))$traits %>%
  select(taxon_name, dataset_id) %>% 
  distinct() %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(dataset_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct(taxon_name, source_id) %>%
  mutate(habitat = "woodland")

habitats <- aquatic %>%
  bind_rows(alpine, bogs_fens, closed_forest, dry_forest, dry_sclerophyll,  eucalypt_forest, grassland, heath, open_forest, rainforest, rainforest2, savanna, shrubland, wet_heath, wet_forest, wet_sclerophyll, woodland) %>%
  group_by(taxon_name) %>%
  mutate(
    source_id = paste0(source_id, collapse = "; ")
  ) %>%
  ungroup() %>%
  distinct() %>%
  group_by(taxon_name, source_id)  %>%
  mutate(
    habitat = paste0(habitat, collapse = " ")
  )%>%
  ungroup() %>%
  distinct()

habitats <- habitats %>%
  tidyr::separate_longer_delim(cols = source_id, delim = "; ") %>%
  distinct() %>%
  arrange(source_id) %>%
  group_by(taxon_name, habitat) %>%
  mutate(source_id = paste0(source_id, collapse = "; ")) %>%
  ungroup() %>%
  distinct() %>%
  tidyr::separate_longer_delim(cols = habitat, delim = " ") %>%
  distinct() %>%
  arrange(habitat) %>%
  group_by(taxon_name, source_id) %>%
  mutate(habitat = paste0(habitat, collapse = " ")) %>%
  ungroup() %>%
  distinct()

habitats %>% write_csv("data/AusTraits_habitats_2024/data.csv")
```

You can add options to executable code like this

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
