---
title: "Threatened species script"
format: html
toc: true
toc-depth: 4
editor: visual
embed-resources: true
editor_options: 
  chunk_output_type: console
---

# Setup and data extraction

## load libraries, build database

- the repo with project-specific data is at https://github.com/traitecoevo/threatened_species
- you can source the most recent austraits.build (AusTraits) release from Zenodo (7.0.0, uploaded 2025-06) or use a data file provided

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(remake)
library(kableExtra)
library(austraits)
library(APCalign)
library(traits.build) # not required if you read in complete RDS file

resources <- APCalign::load_taxonomic_resources(version = "2026-01-14")

options(knitr.kable.NA = "") # blank cells (not literal "NA") for NA values in every kable() table below

source("~/GitHub/threatened_species_traits/R/custom_R_code.R") # not required if you read in complete RDS file
source("~/GitHub/threatened_species_traits/R/extra_functions.R")
source("~/GitHub/austraits.portal/R/species_means.R")

#austraits <- load_austraits(version = "7.0.0") # not required if you read in complete RDS file

#traits.build::build_setup_pipeline(method = "remake", database_name = "threatened_species") # not required if you read in complete RDS file

#threatened_species <- remake::make("threatened_species") # not required if you read in complete RDS file

# create combined database with both data sources

#austraits_with_NVIS <-readRDS("~/GitHub/threatened_species_traits/data_extras/austraits_with_NVIS.rds") # static file with all data 2025-06-16

#all_traits <- austraits_with_NVIS %>% austraits::bind_databases(threatened_species) # not required if you read in complete RDS file

#write_rds(all_traits, "~/GitHub/threatened_species_traits/data_extras/all_traits.rds")

all_traits <- read_rds("~/GitHub/threatened_species_traits/data_extras/all_traits.rds")
```

## create dataframes for key traits

### read in categorical to ordinal score conversions for categorical traits and threats

- read in files that indicate conversions between categorical values and an ordinal equivalent
- these ordinal values are used both to determine if trait-susceptibility correlations exist within training datasets (for myrtle rust, Phytophthora cinnamomi, and goats) and also to assign susceptibility based on a species' trait values

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# categorical trait scores
life_history_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/life_history_scores.csv")
plant_growth_form_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/plant_growth_form_scores.csv")
plant_growth_substrate_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/plant_growth_substrate.csv")
woodiness_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/woodiness_scores.csv")
dispersers_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/dispersers_scores.csv")
vegetative_reproduction_ability_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/vegetative_reproduction_ability_scores.csv")
seedbank_location_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/seedbank_location_scores.csv")
fruit_fleshiness_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/fruit_fleshiness_scores.csv")
bud_bank_location_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/bud_bank_location_scores.csv") %>%
  mutate(bud_bank_location_scaled = bud_bank_location/5)
storage_organ_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/storage_organ_scores.csv")
leaf_glaucousness_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/leaf_glaucousness_scores.csv")
resprouting_capacity_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/resprouting_capacity_scores.csv")
nitrogen_fixing_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/nitrogen_fixing_scores.csv")
leaf_hairs_adult_leaves_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/leaf_hairs_adult_leaves_scores.csv")
leaf_glaucousness_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/leaf_glaucousness_scores.csv")
plant_spinescence_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/plant_spinescence_scores.csv")
plant_physical_defence_structures_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/plant_physical_defence_structures_scores.csv")
pollination_syndrome_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/pollination_syndrome_scores.csv")

# susceptibility attribute scores
plant_goat_toxicity_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/plant_goat_toxicity_scores.csv")
plant_goat_palatability_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/plant_goat_palatability_scores.csv")
phytophthora_cinnamomi_susceptibility_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/phytophthora_cinnamomi_susceptibility_scores.csv")
myrtle_rust_susceptibility_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/myrtle_rust_susceptibility_scores.csv")

# habitat, soil attribute scores - scored differently for each threat
soil_type_scores <- read_csv("~/GitHub/threatened_species_traits/data_extras/soil_type_scores.csv")
geologic_substrate_scores <-  read_csv("~/GitHub/threatened_species_traits/data_extras/geologic_substrate_scores.csv")
habitat_scores <-  read_csv("~/GitHub/threatened_species_traits/data_extras/habitat_scores.csv")
habitat_moisture_scores <-  read_csv("~/GitHub/threatened_species_traits/data_extras/habitat_moisture_scores.csv")
```

### functions

#### extract_numeric_trait

function to extract numeric traits

```{r, echo=FALSE, message=FALSE, warning=FALSE}
extract_numeric_trait <- function(database, numeric_trait) {
  
  extracted_data <- database %>%
    austraits::extract_trait(numeric_trait, partial_matches_allowed = FALSE) %>%
    austraits::extract_data(table = "traits", col = "basis_of_record", col_value = c("field", "literature", "field literature", "literature field", "preserved_specimen"), partial_matches_allowed = F) |>
    austraits::join_location_coordinates() |>
    austraits::join_methods(vars="all") |>
    austraits::join_taxa(vars = "all")
    
  extracted_data$traits <- extracted_data$traits %>%
    dplyr::mutate(
      value = ifelse(str_detect(value, "<"), str_replace(value, "<", "0--"), value),
      value_type = ifelse(value_type == "bin", "minimum--maximum", value_type)
      ) %>%
    separate_longer_delim(cols = c(value_type, value), delim = "--")
  
  extracted_data <- extracted_data$traits %>%
    estimate_species_trait_means() %>%
    filter(taxon_name %in% resources$APC_accepted$canonical_name) %>%
    select(taxon_name, value_mean) %>%
    mutate(
      log10 = log10(as.numeric(value_mean)),
      log10 = ifelse(is.infinite(log10), NA, log10),
      z_score = (log10 - mean(log10, na.rm = TRUE)) /
                sd(log10, na.rm = TRUE),
      scaled = (log10 - min(log10, na.rm = TRUE)) /
                (max(log10, na.rm = TRUE) - min(log10, na.rm = TRUE)),
      # non-linear scaling that will pull extreme outliers in towards 0/1 while spreading out the middle:
      sigmoid_scaled = 1 / (1 + exp(-z_score))) %>% 
    rename(
      !!numeric_trait := value_mean,
      !!paste0(numeric_trait, "_log10") := log10,
      !!paste0(numeric_trait, "_z_score") := z_score,
      !!paste0(numeric_trait, "_scaled") := scaled,
      !!paste0(numeric_trait, "_sigmoid_scaled") := sigmoid_scaled
      )
    
  extracted_data
}
```

#### extract_categorical_trait

function to extract categorical traits

```{r, message=FALSE, warning=FALSE}
extract_categorical_trait <- function(database = database, categorical_trait, trait_scores, output_name) {
  
  extracted_data <- (database %>%
  austraits::extract_trait(categorical_trait, partial_matches_allowed = FALSE))$traits %>%
  tidyr::separate_longer_delim(value, delim = " ") %>%
  dplyr::rename(trait_value = value) %>%
  dplyr::left_join(trait_scores %>% rename(output_name = 3), by = join_by(trait_name, trait_value)) %>%
  dplyr::filter(!is.na(output_name)) %>%
  dplyr::select(dataset_id, observation_id, taxon_name, output_name) %>%
  dplyr::group_by(taxon_name, dataset_id, observation_id) %>%
  dplyr::mutate(output_name = mean(as.numeric(output_name))) %>%
  dplyr::ungroup() %>%
  dplyr::distinct() %>%
  dplyr::select(taxon_name, output_name) %>%
  group_by(taxon_name) %>%
  dplyr::mutate(output_name = mean(as.numeric(output_name))) %>%
  dplyr::ungroup() %>%
  dplyr::distinct() %>%
  rename(!!output_name := output_name)
  
  extracted_data
}
```

#### calculate_prop_traits_scored

```{r, message=FALSE, warning=FALSE}
calculate_prop_traits_scored <- function(data, trait_table) {
  
  traits <- trait_table %>% pull(trait)
  
  data %>%
    rowwise() %>%
      mutate(prop_traits_scored = sum(!is.na(pick(all_of(traits)))) / length(traits)) %>%
    ungroup()
}

calculate_prop_traits_scored_scaled <- function(data, trait_table) {

  traits <- trait_table %>% pull(trait)

  # traits with effect_score = NA are handled manually elsewhere (e.g. pigs'
  # fruit_fleshiness, merged into fruit_length's scoring) and have no defined weight
  # they are still included in this table for display, but are not used in processing
  trait_weights <- trait_table %>%
    dplyr::mutate(effect_score = tidyr::replace_na(abs(effect_score), 0)) %>%
    dplyr::select(trait, effect_score) %>%
    deframe() # Converts to named vector: names = traits, values = weights

  total_effect_scores <- sum(trait_weights)

  # Claude-coded; to speed up processing by 1200x the data are vectorised: 
  # a single matrix multiplication of the (rows x traits) "not NA" 0/1
  # matrix against the weight vector gives every row's weighted trait count in one
  # pass. Replaces the previous rowwise() + sapply() + cur_data() version, which made
  # one function call per trait per row
  not_na <- !is.na(as.matrix(dplyr::select(data, all_of(traits))))
  weighted_traits_scored <- as.numeric(not_na %*% trait_weights[traits]) / total_effect_scores

  data %>%
    mutate(prop_traits_scored = weighted_traits_scored)
}
```

#### get_effect_score, calculate_susceptibility and add_trait_susceptibility

- `get_effect_score` gets effect scores from the table for each threat; for some threat x trait combinations need to manually input value because of additional conditions

- `calculate_susceptibility` scales all ordinal or scaled numeric values for a trait or taxonomic susceptibility, based on effect score and ordinal value

- `add_trait_susceptibility` calls `get_effect_score` + `calculate_susceptibility` for every trait in a threat's trait_effects table in one go, replacing the block of near-identical `{trait}_susceptibility = calculate_susceptibility(...)` lines that used to be repeated (one per trait) in every threat section

```{r, message=FALSE, warning=FALSE}
get_effect_score <- function(data, trait_name) {
  data %>%
    filter(trait == trait_name) %>%
    slice(1) %>%  # just in case there are duplicates
    as.list()
}

calculate_susceptibility <- function(trait, effect_score, data) {
  ifelse(!is.na(data[[trait]]),
         effect_score * (data[[trait]] - mean(data[[trait]], na.rm = TRUE)),
         NA
  )
}

# `add_trait_susceptibility` replaces the repeated block of
#   {trait}_susceptibility = calculate_susceptibility("{trait}", get_effect_score(trait_effects, "{trait}")$effect_score, .)
# lines that were copy-pasted (one line per trait) in every threat section. It loops
# over every trait in `trait_effects` that has a non-NA effect_score (traits with
# effect_score = NA are handled manually elsewhere, e.g. merged into another trait's
# scoring, so are deliberately skipped here) and adds a `{trait}_susceptibility` column for each.
#
# `condition`: optional logical vector (same length/order as `data`'s rows), for
# threats where trait susceptibility only applies to a subset of taxa - e.g. myrtle rust traits
# only apply to Myrtaceae. Defaults to TRUE (applies to all rows). Rows where the
# condition is FALSE get NA, matching the old per-trait `ifelse(family == "Myrtaceae", calculate_susceptibility(...), NA)` pattern.
add_trait_susceptibility <- function(data, trait_effects, condition = TRUE) {

  # recycle condition to nrow(data) BEFORE the loop - if left as the scalar default
  # TRUE, ifelse(condition, ...) below would take its length from `condition` (length
  # 1), silently truncating every susceptibility column to its first value and recycling that
  # one number across all rows. (Caught by testing against the old manual code with
  # synthetic data before this was applied to the real pipeline - see chat history.)
  condition <- rep_len(condition, nrow(data))

  traits_to_score <- trait_effects %>% filter(!is.na(effect_score)) %>% pull(trait)

  for (trait in traits_to_score) {
    susceptibility_col <- paste0(trait, "_susceptibility")
    trait_effect <- get_effect_score(trait_effects, trait)$effect_score
    data[[susceptibility_col]] <- ifelse(condition, calculate_susceptibility(trait, trait_effect, data), NA)
  }

  data
}
```

#### make_8_panel_susceptibility_density_plot

function to plot susceptibility densities

```{r, echo=FALSE, message=FALSE, warning=FALSE}
make_8_panel_susceptibility_density_plot <- function(data, threat) {

  get_legend <- function(myplot) {
    tmp <- ggplotGrob(myplot)
    legend_index <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
    tmp$grobs[[legend_index]]
  }

  susceptibility_colors <- c(
    "habitat susceptibility" = "#FFFF33",
    "trait and taxonomy susceptibility" = "#F46D43",
    "combined individual susceptibility" = "#74ADD1",
    "min population susceptibility" = "#4DAF4A",
    "max population susceptibility" = "#984EA3",
    "mean population susceptibility" = "#E7298A"
  )

  data$category <- factor(data$category, levels = c(
    "habitat susceptibility", "trait and taxonomy susceptibility", "combined individual susceptibility", "min population susceptibility", "max population susceptibility",
    "mean population susceptibility"
  ))

  data_threatened <- data  %>%
    filter(!is.na(threatened_status))

  # the min/max row is only added when the parameters are specified
  has_min_max <- any(c("min susceptibility", "max susceptibility") %in% data_threatened$category)

  # the population-susceptibility row (this threat's habitat x trait score merged with regeneration
  # ability & precariousness, then rescaled to population_susceptibility) is only added
  # once a threat's calc chunk has been extended with that merge - most threats haven't
  # been yet, so this falls back to not showing the row rather than erroring. Checked
  # against the full `data` (not data_threatened), since - unlike min/max, which is
  # listed-taxa only - this row shows both populations, like row1/row2.
  has_population_susceptibility <- any(data$category == "mean population susceptibility")

  # shared panel styling, previously copy-pasted once per plot (plot1-4) - now built once
  # and reused for every panel, including the new min/max ones
  density_panel <- function(plot_data) {
    ggplot(plot_data, aes(x = susceptibility, fill = category)) +
      geom_density(aes(y = ..scaled..), alpha = 0.4) +
      scale_fill_manual(values = susceptibility_colors, name = "category") +
      scale_x_continuous(breaks = seq(0, 1, by = 0.2)) +
      # coord_cartesian() (not scale_x_continuous(limits=...)) crops the *view* only -
      # the density curve is still estimated from the full data first, then the window
      # is cropped. scale_x_continuous(limits=...) instead silently drops any row outside
      # the range BEFORE estimating the density, which badly distorted the min/max susceptibility
      # panels - up to 69% of a threat's total_individual_susceptibility_min values fall below -0.1, since
      # margins can push values well past the range the original 3 categories ever used.
      # This is a fixed window (not adaptive) so every threat plots on the same, directly
      # comparable x-axis - if a rescale range (linear or sigmoidal_rescale) pushes total_individual_susceptibility
      # outside c(-.1, 1.5), that's a cue to adjust the scaling range itself, not the plot.
      coord_cartesian(xlim = c(-0.2, 1.5)) +
      labs(x = "", y = "", fill = NULL) +
      theme_minimal()  +
      theme(
        panel.grid.major.x = element_line(color = "gray80"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none"
      )
  }

  dummy_plot <- ggplot(data, aes(x = susceptibility, fill = category)) +
    geom_density(aes(y = ..scaled..), alpha = 0.4) +
    scale_fill_manual(values = susceptibility_colors, name = "category")  +
    theme(legend.position = "bottom")

  legend <- get_legend(dummy_plot)

#  row 1, individual susceptibilities based on traits and habitat
  plot1 <- density_panel(data %>% filter(category %in% c("habitat susceptibility", "trait and taxonomy susceptibility")))
  plot2 <- density_panel(data_threatened %>% filter(category %in% c("habitat susceptibility", "trait and taxonomy susceptibility")))
                         
  title_row1 <- grid::textGrob("Individual susceptibility components",
                               x = 0.07, hjust = 0.0,
                               gp = grid::gpar(fontsize = 10))
  row1 <- gridExtra::arrangeGrob(plot1, plot2, nrow = 1, ncol = 2, top = title_row1)

# row 2, combined individual susceptibility
  plot3 <- density_panel(data %>% filter(category == "combined individual susceptibility"))
  plot4 <- density_panel(data_threatened %>% filter(category == "combined individual susceptibility"))
  
  title_row2 <- grid::textGrob("Total individual susceptibility (traits, taxonomy, habitat combined)",
                               x = 0.07, hjust = 0.0,
                               gp = grid::gpar(fontsize = 10))
  row2 <- gridExtra::arrangeGrob(plot3, plot4, nrow = 1, ncol = 2, top = title_row2)

# row 3, mean population susceptibility
  plot5 <- density_panel(data %>% filter(category == "mean population susceptibility"))
  plot6 <- density_panel(data_threatened %>% filter(category == "mean population susceptibility"))

  title_row3 <- grid::textGrob(
      "Population susceptibility, including individual susceptibility, population regeneration ability & population precariousness",
      x = 0.07, hjust = 0.0,
      gp = grid::gpar(fontsize = 10))

  row3 <- gridExtra::arrangeGrob(plot5, plot6, nrow = 1, ncol = 2, top = title_row3)

  # row 4, min/max population susceptibility
  plot7 <- density_panel(data_threatened %>% filter(category == "min population susceptibility"))
  plot8 <- density_panel(data_threatened %>% filter(category == "max population susceptibility"))

  title_row_4 <- grid::textGrob("Range of population susceptibilities (listed species only)",
                                x = 0.07, hjust = 0.0,
                                gp = grid::gpar(fontsize = 10))

  row4 <- gridExtra::arrangeGrob(plot7, plot8, nrow = 1, ncol = 2, top = title_row_4)

  # population header - "All species" / "Listed species" - previously squeezed in as a
  # single manually-spaced string under row 3; every row splits the same way (all taxa
  # left, listed taxa right), so it only needs to be said once, in a larger font, right
  # under the overall title rather than buried mid-plot
  title_population_all <- grid::textGrob("All species",
                                gp = grid::gpar(fontsize = 14, fontface = "bold"))
  title_population_listed <- grid::textGrob("Listed species",
                                gp = grid::gpar(fontsize = 14, fontface = "bold"))

  row0 <- gridExtra::arrangeGrob(title_population_all, title_population_listed, nrow = 1, ncol = 2)

combined <- gridExtra::arrangeGrob(grobs = list(row0, row1, row2, row3, row4),
                                   nrow = 5, ncol = 1, heights = c(0.25, 1, 1, 1.2, 1), top = paste0("Distribution of Threat-dependent Susceptibility Scores for ", threat))

  combined_with_legend <- gridExtra::arrangeGrob(combined, legend, nrow = 2, heights = c(0.9, 0.1))

  grid::grid.draw(combined_with_legend)

  # invisible, not return() - the plot is drawn above via grid.draw(); returning visibly
  # makes this grob the chunk's last value, which knitr then auto-prints as literal
  # "TableGrob (...): N grobs" text under every plot
  invisible(combined_with_legend)
}
```

#### make_2_panel_susceptibility_density_plot

Plotting function for threat-independent effects, where no habitat value and no confidence scores.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
make_2_panel_susceptibility_density_plot <- function(data, threat) {

  get_legend <- function(myplot) {
    tmp <- ggplotGrob(myplot)
    legend_index <- which(sapply(tmp$grobs, function(x) x$name) == "guide-box")
    tmp$grobs[[legend_index]]
  }

  susceptibility_colors <- c(
    "trait susceptibility" = "#F46D43"
  )

  data$category <- factor(data$category, levels = c("trait susceptibility"))

  data_threatened <- data %>%
    filter(!is.na(threatened_status))

  # shared panel styling - same fixed x-axis window as make_8_panel_susceptibility_density_plot() so every
  # plot in the report is directly comparable
  density_panel <- function(plot_data) {
    ggplot(plot_data, aes(x = susceptibility, fill = category)) +
      geom_density(aes(y = ..scaled..), alpha = 0.4) +
      scale_fill_manual(values = susceptibility_colors, name = "category") +
      scale_x_continuous(breaks = seq(0, 0.15, by = 0.05)) +
      coord_cartesian(xlim = c(-0.01, 0.15)) +
      labs(x = "", y = "", fill = NULL) +
      theme_minimal() +
      theme(
        panel.grid.major.x = element_line(color = "gray80"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none"
      )
  }

  dummy_plot <- ggplot(data, aes(x = susceptibility, fill = category)) +
    geom_density(aes(y = ..scaled..), alpha = 0.4) +
    scale_fill_manual(values = susceptibility_colors, name = "category") +
    theme(legend.position = "bottom")

  legend <- get_legend(dummy_plot)

  plot1 <- density_panel(data %>% filter(category == "trait susceptibility"))
  plot2 <- density_panel(data_threatened %>% filter(category == "trait susceptibility"))

  title_all <- grid::textGrob(
    "susceptibility scores for all Australian plant taxa",
    x = 0.07, hjust = 0.1,
    gp = grid::gpar(fontsize = 10))

  title_listed <- grid::textGrob(
    "susceptibility scores for listed plant taxa",
    x = 0.07, hjust = 0.1,
    gp = grid::gpar(fontsize = 10))

  title_x <- grid::textGrob(
    "susceptibility",
    gp = grid::gpar(fontsize = 10))

  # row 1's two panels differ by population (all taxa vs listed taxa), not by category,
  # so each gets its own title rather than one shared row title
  row1 <- gridExtra::arrangeGrob(
    gridExtra::arrangeGrob(plot1, top = title_all),
    gridExtra::arrangeGrob(plot2, top = title_listed),
    nrow = 1, ncol = 2
  )

  combined <- gridExtra::arrangeGrob(row1, nrow = 1, ncol = 1,
                  top = paste0("Distribution of Trait-based Susceptibility Scores for ", threat))

  combined_with_legend <- gridExtra::arrangeGrob(combined, legend, nrow = 1, widths = c(0.8, 0.2))

  grid::grid.draw(combined_with_legend)

  # invisible, not return() - see make_8_panel_susceptibility_density_plot() for why
  invisible(combined_with_legend)

}
```

#### make_4_panel_susceptibility_density_plot

Plotting function for the report figure used in every threat's Word report ("Distribution of {habitat & taxonomy susceptibility / trait susceptibility / individual susceptibility / population susceptibility} for currently-listed {threat} taxa"). Reads directly from each threat's own `_assigned_susceptibility` dataframe - the same object the calc chunk itself produces - so this is guaranteed to match the pipeline's real numbers rather than a re-derivation from the exported CSV. Two- or three-panel threats (e.g. Phytophthora's `trait & taxonomy susceptibility` is a single combined column, not separate `habitat` and `trait` columns) can just supply fewer panels via `panel_cols`/`panel_labels`/`panel_colors`. `x_min`/`x_max` set the shared x-axis window across all panels (default -0.2 to 2.1, matching population_susceptibility's widest plausible range) - narrow this for threats whose scores stay well under that ceiling.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
make_4_panel_susceptibility_density_plot <- function(data, threat,
                                            panel_cols,
                                            panel_labels = panel_cols,
                                            panel_colors = c("#FFFF33", "#F46D43", "#74ADD1", "#E7298A")[seq_along(panel_cols)],
                                            listed_only = TRUE,
                                            x_min = -0.2,
                                            x_max = 2.1) {

  stopifnot(length(panel_cols) == length(panel_labels), length(panel_cols) <= length(panel_colors))

  plot_data <- data
  if (listed_only) plot_data <- plot_data %>% filter(!is.na(threatened_status))

  df_long <- plot_data %>%
    select(all_of(panel_cols)) %>%
    setNames(panel_labels) %>%
    tidyr::pivot_longer(everything(), names_to = "category", values_to = "susceptibility") %>%
    mutate(category = factor(category, levels = panel_labels))

  susceptibility_colors <- setNames(panel_colors[seq_along(panel_labels)], panel_labels)

  density_panel <- function(one_category) {
    ggplot(df_long %>% filter(category == one_category), aes(x = susceptibility, fill = category)) +
      geom_density(aes(y = ..scaled..), alpha = 0.5) +
      scale_fill_manual(values = susceptibility_colors, name = "category") +
      coord_cartesian(xlim = c(x_min, x_max)) +
      labs(x = "", y = "", fill = NULL, title = one_category) +
      theme_minimal(base_size = 13) +
      theme(
        panel.grid.major.x = element_line(color = "gray80"),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        axis.text.y = element_blank(),
        legend.position = "none",
        plot.title = element_text(face = "bold", size = 11)
      )
  }

  panels <- lapply(panel_labels, density_panel)
  n <- length(panels)
  ncol <- if (n <= 2) n else 2

  combined <- gridExtra::arrangeGrob(
    grobs = panels, ncol = ncol,
    top = paste0("Distribution of ", threat, " susceptibility scores for currently-listed plant taxa")
  )

  grid::grid.draw(combined)

  # invisible, not return() - see make_8_panel_susceptibility_density_plot() for why
  invisible(combined)
}
```

#### create_susceptibility_bins

turn continuous susceptibility or susceptibility scores into ordered categorical bins from a named threshold vector.

`bin_thresholds` defaults to the standard scheme used by most threats (0.8/0.6/0.4/0.2); pass a different named vector for threats that use different cutoffs (e.g. C4 grasses uses 0.7/0.4/0.3/0.15). This is the ONLY place the standard scheme's values should be defined - every threat section defers to this default rather than keeping its own copy, so there's a single source of truth.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
create_susceptibility_bins <- function(x, bin_thresholds = c(
      "5_extremely_high" = 0.8,
      "4_high"           = 0.6,
      "3_moderate"       = 0.4,
      "2_low"            = 0.2,
      "1_no_susceptibility"        = -10 # well below 0, in case scaling causes negative values
    )) {
  # bin_thresholds must be sorted descending by value, with names = bin labels
  labels  <- names(bin_thresholds)
  cutoffs <- unname(bin_thresholds)

  ord <- order(cutoffs, decreasing = TRUE)
  labels  <- labels[ord]
  cutoffs <- cutoffs[ord]

  result <- rep(NA_character_, length(x))
  for (i in seq_along(cutoffs)) {
    result[is.na(result) & x >= cutoffs[i]] <- labels[i]
  }
  result
}
```

#### add_susceptibility_range

Formalises turning susceptibility_confidence into a plausible range around a numeric susceptibility score, instead of the point estimate alone. Low confidence -\> wide range; high confidence -\> narrow range.

margin = (1 - confidence) / divisor {output_base}\_min = susceptibility - margin {output_base}\_max = susceptibility + margin {output_base}\_bin_min = create_susceptibility_bins(min) {output_base}\_bin_max = create_susceptibility_bins(max)

divisor defaults to 5 (i.e. at confidence 0, margin is 1/5 = 0.2 either side, rather than the full 0-1 range; at confidence 0.6 - roughly the top of the observed range - margin is 0.4/5 = 0.08) - change it here (or per-call) and every threat regenerates.

susceptibility_col/confidence_col: names (strings) of the numeric susceptibility and confidence columns to use as inputs. output_base: name to build the four new columns from - defaults to susceptibility_col, but can differ, e.g. for Phytophthora/myrtle rust, where the numeric value that actually determines population_susceptibility_bin is total_individual_susceptibility_overwrite (not total_individual_susceptibility), so susceptibility_col = "total_individual_susceptibility_overwrite" but output_base = "total_individual_susceptibility" keeps the output column names (total_individual_susceptibility_min/max/score_min/score_max) consistent with every other threat. bin_thresholds: passed through to create_susceptibility_bins(); NULL uses its own default.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

add_susceptibility_range <- function(data, susceptibility_col, confidence_col, divisor = 5, output_base = susceptibility_col, bin_thresholds = NULL) {

  min_col <- paste0(output_base, "_min")
  max_col <- paste0(output_base, "_max")
  score_min_col <- paste0(output_base, "_bin_min")
  score_max_col <- paste0(output_base, "_bin_max")

  margin <- (1 - data[[confidence_col]]) / divisor
  data[[min_col]] <- data[[susceptibility_col]] - margin
  data[[max_col]] <- data[[susceptibility_col]] + margin

  if (is.null(bin_thresholds)) {
    data[[score_min_col]] <- create_susceptibility_bins(data[[min_col]])
    data[[score_max_col]] <- create_susceptibility_bins(data[[max_col]])
  } else {
    data[[score_min_col]] <- create_susceptibility_bins(data[[min_col]], bin_thresholds)
    data[[score_max_col]] <- create_susceptibility_bins(data[[max_col]], bin_thresholds)
  }

  data
}
```

#### bin_thresholds_table

Turns a `bin_thresholds` named vector (as used by `create_susceptibility_bins()`) into a tibble of `population_susceptibility_bin` + a human-readable `range` string, for joining onto a `count(population_susceptibility_bin)` summary table.

Each bin's lower bound is its own threshold; its upper bound is the threshold of the bin above it. The top bin is left open-ended above ("from x"), the bottom bin open-ended below ("to x"), and every other bin is printed as "lower–upper" (en dash). Values are printed as-is - no rounding.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
bin_thresholds_table <- function(bin_thresholds) {
  enframe(bin_thresholds, name = "population_susceptibility_bin", value = "threshold") %>%
    arrange(desc(threshold)) %>%
    mutate(
      upper = lag(threshold),
      range = case_when(
        is.na(upper)      ~ paste0(threshold, "–1.0"),
        TRUE              ~ paste0(threshold, "–", upper)
      )
    ) %>%
    select(population_susceptibility_bin, range)
}
```

#### sigmoidal_rescale

scales::rescale() is a pure linear min-max stretch: whichever single species has the most extreme total_trait_effect_score gets pinned exactly to the top and bottom boundaries, and other species' positions are relative to that one point. A single outlier can strongly influence the distibution.

sigmoidal_rescale() centers on the MEAN and SD of `total_trait_effect_score` instead of its min/max, then lets values fall away from center via a sigmoid rather than a linear ramp - so extreme species approach edges of the distribution smoothly

`k` is the spread knob: bigger k = gentler compression = fatter, more spread-out tails, same center regardless. Start at k = 1 and adjust.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

sigmoidal_rescale_k <- 1

sigmoidal_rescale <- function(x, to = c(0, 1), center = mean(x, na.rm = TRUE),
                          spread = sd(x, na.rm = TRUE), k = sigmoidal_rescale_k) {
  z <- (x - center) / (spread * k)
  s <- 1 / (1 + exp(-z))          # sigmoid: s(0) = 0.5 exactly at the center, asymptotes to 0/1
  to[1] + s * (to[2] - to[1])
}

```

#### check_trait_correlations

Runs a univariate lm(response_var \~ trait) for every trait in `vars`, collects r2/p-value/coefficient/n into a results tibble, and plots (point + jitter + lm smooth) any trait that clears both `r2_threshold` and `p_threshold`. Can be used for catgorical or numeric traits; for numeric traits use a log10 transform via `use_log10 = TRUE`.

```{r, echo=FALSE, message=FALSE, warning=FALSE}

check_trait_correlations <- function(data, response_var, vars, r2_threshold, p_threshold = 0.05, use_log10 = FALSE) {

  results <- tibble(trait = character(), coeff = character(), r2 = character(), p_value = character(), n = character())

  for (var in vars) {
    data_tmp <- data %>% filter(!is.na(.data[[var]]), !is.na(.data[[response_var]]))
    data_tmp$predictor <- if (use_log10) log10(as.numeric(data_tmp[[var]])) else data_tmp[[var]]

    linearmodel_var <- lm(data_tmp[[response_var]] ~ predictor, data = data_tmp)

    p     <- summary(linearmodel_var)[4] %>% as.data.frame() %>% select(4)
    coeff <- summary(linearmodel_var)[4] %>% as.data.frame() %>% select(1)

    new_row <- c(
      trait   = var,
      r2      = round(summary(linearmodel_var)$r.squared, 4),
      p_value = round(p[2, 1], 4),
      coeff   = round(coeff[2, 1], 3),
      n       = nrow(data_tmp)
    ) %>% t() %>% as.data.frame()

    if (new_row[1, ]$r2 > r2_threshold & new_row[1, ]$p_value < p_threshold) {
      plot(
        ggplot(data_tmp, aes(x = predictor, y = data_tmp[[response_var]])) +
          geom_point() +
          geom_jitter() +
          geom_smooth(method = "lm", se = FALSE, color = "grey20") +
          theme_classic() +
          xlab(var)
      )
    }

    results <- results %>% bind_rows(new_row)
  }

  results
}
```

### categorical-ordinal score conversions

Generate tables of ordinal score conversions for categorical traits and attributes.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#print out tables for different categorical traits, showing ordinal scores

plant_growth_form_scores %>% arrange(trait_value) %>% kable(caption = "plant growth form values, with different ordinal value options") %>% kable_classic()

plant_growth_substrate_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value) %>% kable(caption = "plant growth substrate values") %>% kable_classic()

woodiness_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value) %>% kable(caption = "plant woodiness values") %>% kable_classic()

resprouting_capacity_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value) %>% kable(caption = "resprouting capacity values") %>% kable_classic()

storage_organ_scores %>% arrange(storage_organ_present) %>% kable(caption = "storage organ values, with different ordinal value options") %>% kable_classic()

bud_bank_location_scores %>% arrange(bud_bank_location) %>% kable(caption = "bud bank location scaled") %>% kable_classic()

nitrogen_fixing_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value) %>% kable(caption = "nitrogen fixing values") %>% kable_classic()

leaf_glaucousness_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value)  %>% kable(caption = "leaf glaucousness values") %>% kable_classic()

leaf_hairs_adult_leaves_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value) %>% kable(caption = "leaf hair values") %>% kable_classic()

plant_spinescence_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value)  %>% kable(caption = "plant spinescence values") %>% kable_classic()

plant_physical_defence_structures_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value)  %>% kable(caption = "plant physical defence structures") %>% kable_classic()

dispersers_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value) %>% kable(caption = "dispersal trait values") %>% kable_classic()

pollination_syndrome_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value) %>% kable(caption = "pollination syndrome values") %>% kable_classic()

vegetative_reproduction_ability_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value) %>% kable(caption = "vegetative reproduction ability values") %>% kable_classic()

seedbank_location_scores %>% rename(ordinal_value = !!names(.)[3]) %>% arrange(ordinal_value) %>% kable(caption = "seedbank location values") %>% kable_classic()
```

### extract numeric traits from database

- extract all available values for numeric traits identified as 1) having sufficient coverage **AND** 2) being plausible relevant to predict the susceptibility of one of the threats

- for each taxon, take the mean of all available values using a weighted mean function developed for AusTraits, which first takes the mean of multiple individual values collected at a single site; then the mean of values from all population; and at the end also considers values extracted from floras (and similar publications) that refer to the entire species

- the function used is called `austraits_weighted_means`

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
numeric_traits <- c("seed_dry_mass", "leaf_mass_per_area", "plant_height", "leaf_length", "leaf_mass_per_area", "leaf_N_per_dry_mass", "leaf_P_per_dry_mass", "reproductive_maturity", "seedbank_longevity", "leaf_water_content_per_area", "leaf_tannin_per_dry_mass", "leaf_phenol_per_dry_mass", "leaf_lignin_per_dry_mass", "digestible_mass_per_dry_mass", "leaf_crude_protein_per_dry_mass", "leaf_work_to_shear_adjusted", "fruit_length", "metabolisable_energy", "bark_thickness", "lifespan", "extent_of_occurrence", "area_of_occupancy", "geographic_range")

traits2 <- c("leaf_soluble_protein_per_area", "leaf_insoluble_protein_per_area")  # interesting traits but only data for 1-2 species - in a lab study

numeric_data <- list()

for (trait in numeric_traits) {
  
  numeric_data[[trait]] <- extract_numeric_trait(all_traits, trait)
  numeric_data[[trait]] 
  
}

list2env(numeric_data, envir = .GlobalEnv)
```

#### add special plant height column to capture impact of tall herbivores

Extra step of creating a special plant height columns, which maxes out at 4 m. This has been designated the plant height more damaged by goats/deer,as they prefer browsing over grazing, but taller plants do better again, assuming they successfully establish

```{r, results='hide', message=FALSE, warning=FALSE}

plant_height <- plant_height %>%
  mutate(
    plant_height_above_4m_log10 = ifelse(plant_height > 4, plant_height_log10, NA),
    plant_height_below_4m_log10 = ifelse(plant_height <= 4, plant_height_log10, NA),
    plant_height_above_4m_scaled = 1 - (plant_height_above_4m_log10 - min(plant_height_above_4m_log10, na.rm = TRUE)) /
              (max(plant_height_above_4m_log10, na.rm = TRUE) - min(plant_height_above_4m_log10, na.rm = TRUE)),
    plant_height_below_4m_scaled = (plant_height_below_4m_log10 - min(plant_height_below_4m_log10, na.rm = TRUE)) /
              (max(plant_height_below_4m_log10, na.rm = TRUE) - min(plant_height_below_4m_log10, na.rm = TRUE)),
    plant_height_4m_peak_scaled = ifelse(!is.na(plant_height_above_4m_scaled), plant_height_above_4m_scaled, plant_height_below_4m_scaled)
    )
```

#### special lifespan calculations because mostly bin data

lifespan cannot directly be extracted using `extract_numeric_trait()` because most lifespan records are bin/range values (e.g. "\<5"), which need to be converted to minimum and/or maximum values

```{r, results='hide', message=FALSE, warning=FALSE}
lifespan_data <- all_traits %>% extract_trait("lifespan", partial_matches_allowed = FALSE)

lifespan_data$traits <- lifespan_data$traits %>%
  mutate(
    value = ifelse(str_detect(value, "<"), str_replace(value, "<", "0--"), value),
    value_type = ifelse(value_type == "bin", "minimum--maximum", value_type)
    ) %>%
  separate_longer_delim(cols = c(value_type, value), delim = "--")

lifespan <- extract_numeric_trait(lifespan_data, numeric_trait = "lifespan")
```

### extract categorical traits from database

- extract all available values for categorical traits identified as 1) having sufficient coverage **AND** 2) being plausible relevant to predict the susceptibility of one of the threats
- for each trait, merge in numeric conversions
- for each taxon, take the mean of all available values

special cases to note: - `life_history` comes exclusively from two specific compilations, `Wenk_2023` and `Wenk_2025` which together provide near-complete coverage for the Australian flora - `woodiness_detailed` comes exclusively from two specific compilations, `Wenk_2022` and `Wenk_2025` which together provide near-complete coverage for the Australian flora - `plant_growth_form` comes exclusively from two specific compilations, `Wenk_2022` and `Wenk_2025` which together provide near-complete coverage for the Australian flora

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
dispersers <-  extract_categorical_trait(all_traits, c("dispersal_syndrome", "dispersers"), dispersers_scores, "dispersers")
resprouting_capacity <- extract_categorical_trait(all_traits, "resprouting_capacity", resprouting_capacity_scores, "resprouting_capacity")
nitrogen_fixing <- extract_categorical_trait(all_traits, "nitrogen_fixing", nitrogen_fixing_scores, "nitrogen_fixing")
life_history <- extract_categorical_trait(all_traits %>% extract_dataset(c("Wenk_2023", "Wenk_2025")), "life_history", life_history_scores, "life_history")
plant_growth_substrate_hymenanche <- extract_categorical_trait(all_traits, "plant_growth_substrate", plant_growth_substrate_scores %>% select(trait_name, trait_value, plant_growth_substrate = plant_growth_substrate_hymenanche), "plant_growth_substrate_hymenanche")
plant_growth_substrate_pigs <- extract_categorical_trait(all_traits, "plant_growth_substrate", plant_growth_substrate_scores %>% select(trait_name, trait_value, plant_growth_substrate = plant_growth_substrate_pigs), "plant_growth_substrate_pigs")
woodiness <- extract_categorical_trait(all_traits %>% extract_dataset(c("Wenk_2022", "Wenk_2025")), "woodiness_detailed", woodiness_scores, "woodiness")
leaf_hairs_adult_leaves <- extract_categorical_trait(all_traits, "leaf_hairs_adult_leaves", leaf_hairs_adult_leaves_scores, "leaf_hairs_adult_leaves")
leaf_glaucousness <- extract_categorical_trait(all_traits, "leaf_glaucousness", leaf_glaucousness_scores, "leaf_glaucousness")
bud_bank_location_scaled <- extract_categorical_trait(all_traits, "bud_bank_location", bud_bank_location_scores %>% select(trait_name, trait_value, bud_bank_location = bud_bank_location_scaled), "bud_bank_location_scaled")
fruit_fleshiness <- extract_categorical_trait(all_traits, "fruit_fleshiness", fruit_fleshiness_scores, "fruit_fleshiness")
plant_spinescence <- extract_categorical_trait(all_traits, "plant_spinescence", plant_spinescence_scores, "plant_spinescence")
plant_physical_defence_structures <- extract_categorical_trait(all_traits, "plant_physical_defence_structures", plant_physical_defence_structures_scores, "plant_physical_defence_structures")
vegetative_reproduction_ability <- extract_categorical_trait(all_traits, "vegetative_reproduction_ability", vegetative_reproduction_ability_scores, "vegetative_reproduction_ability")
seedbank_location <- extract_categorical_trait(all_traits, "seedbank_location", seedbank_location_scores, "seedbank_location")
pollination_syndrome <-  extract_categorical_trait(all_traits, c("pollination_system", "pollination_syndrome"), pollination_syndrome_scores, "pollination_syndrome")

# various ways to score growth forms based on threat
plant_growth_form <- extract_categorical_trait(all_traits %>% extract_dataset(c("Wenk_2022", "Wenk_2025")), "plant_growth_form", plant_growth_form_scores %>% select(trait_name, trait_value, plant_growth_form), "plant_growth_form")
forb_graminoid <- extract_categorical_trait(all_traits %>% extract_dataset(c("Wenk_2022", "Wenk_2025")), "plant_growth_form", plant_growth_form_scores %>% select(trait_name, trait_value, plant_growth_form = forb_graminoid), "forb_graminoid")
herb_shrub_tree <- extract_categorical_trait(all_traits %>% extract_dataset(c("Wenk_2022", "Wenk_2025")), "plant_growth_form", plant_growth_form_scores %>% select(trait_name, trait_value, plant_growth_form = herb_shrub_tree), "herb_shrub_tree")
geophyte <- extract_categorical_trait(all_traits %>% extract_dataset(c("Wenk_2022", "Wenk_2025")), "plant_growth_form", plant_growth_form_scores %>% select(trait_name, trait_value, plant_growth_form = geophyte), "geophyte")
regeneration <- extract_categorical_trait(all_traits %>% extract_dataset(c("Wenk_2022", "Wenk_2025")), "plant_growth_form", plant_growth_form_scores %>% select(trait_name, trait_value, plant_growth_form = regeneration), "regeneration")

# various ways to score storage organs based on threat
storage_organ_present <- extract_categorical_trait(all_traits, "storage_organ", storage_organ_scores %>% select(trait_name, trait_value, storage_organ = storage_organ_present), "storage_organ_present")

storage_organ_fleshy <- extract_categorical_trait(all_traits, "storage_organ", storage_organ_scores %>% select(trait_name, trait_value, storage_organ = storage_organ_fleshy), "storage_organ_fleshy")

# flowering & fruiting season scorings
tropical_grass_flowering_times <- all_traits$traits %>%
  filter(trait_name == "flowering_time") %>%
  select(taxon_name, value) %>%
  mutate(
    value = stringr::str_replace_all(value, "y", "1"),
    value = stringr::str_replace_all(value, "n", "0")
    ) %>%
  tidyr::separate(value, into = paste0("month", 01:12), sep = 1:11) %>%
  mutate(
    across(month1:month12, as.numeric),
    all_months = rowSums(across(c(month1:month12))),
    wet_season_flowering = rowSums(across(c(month9:month11))) / all_months,
    dry_season_flowering = rowSums(across(c(month12, month1:month4))) / all_months,
    shoulder_flowering = rowSums(across(c(month5:month8))) / all_months
  ) %>%
  select(taxon_name, wet_season_flowering, dry_season_flowering, shoulder_flowering) %>%
  group_by(taxon_name) %>%
    mutate(across(c(1:3), ~mean(.x))) %>%
  ungroup() %>%
  distinct()
```

### extract known susceptibility scores

- these include:

  1.  taxa explicitly known to be susceptible / not susceptible to a specific disease
  2.  taxa explicitly know to be palatable, toxic, etc. to a specific herbivore

special goat palatability notes: - sometimes `plant_goat_palatability` is scored for the entire genus and the value is mapped to all members of the genus

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# extract known susceptibility data
myrtle_rust_susceptibility <- extract_categorical_trait(all_traits, "myrtle_rust_susceptibility", myrtle_rust_susceptibility_scores, "myrtle_rust_susceptibility")

phytophthora_cinnamomi_susceptibility <- extract_categorical_trait(all_traits, "phytophthora_cinnamomi_susceptibility", phytophthora_cinnamomi_susceptibility_scores, "phytophthora_cinnamomi_susceptibility")

# extract known toxicity/palatability data
plant_goat_toxicity <- extract_categorical_trait(all_traits, "plant_goat_toxicity", plant_goat_toxicity_scores, "plant_goat_toxicity")
plant_goat_palatability <- extract_categorical_trait(all_traits, "plant_goat_palatability", plant_goat_palatability_scores, "plant_goat_palatability")

# map genus-level values to individual species
plant_goat_palatability_genera <- plant_goat_palatability %>%
  filter(!str_detect(taxon_name, " ")) %>%
  rename(genus = taxon_name) %>%
  left_join(resources$APC %>% filter(taxonomic_status == "accepted", taxon_rank == "species") %>% select(genus, taxon_name = canonical_name))

plant_goat_palatability <- plant_goat_palatability %>%
  filter(str_detect(taxon_name, " ")) %>%
  bind_rows(plant_goat_palatability_genera)

grasses_impacted_species_tmp <- read_csv("~/GitHub/threatened_species_traits/data_extras/grasses_impacted_species_per_TAP.csv")

grasses_impacted_species <- APCalign::create_taxonomic_update_lookup(grasses_impacted_species_tmp$taxon_name, resources = resources, full = T, fuzzy_matches = F) |>
  select(accepted_name, suggested_name, original_name, everything()) |>
  mutate(taxon_name = ifelse(is.na(accepted_name), original_name, accepted_name)) |>
  select(taxon_name, original_name) |>
  left_join(grasses_impacted_species_tmp, by = c("original_name" = "taxon_name")) |>
  select(-original_name)

# hymenachne: species whose TAP-listed grass_threat explicitly names hymenachne
hymenachne_known_impact_species <- grasses_impacted_species |>
  filter(str_detect(str_to_lower(grass_threat), "hymenachne")) |>
  pull(taxon_name)

# C4 grasses: species whose TAP-listed grass_threat names gamba grass and/or mission
# grass(es), or that don't specify which grass (grass_threat == "-").
c4_grasses_known_impact_species <- grasses_impacted_species |>
  filter(str_detect(str_to_lower(grass_threat), "gamba|mission") | grass_threat == "-") |>
  pull(taxon_name)
```


### extract non-trait attributes from database

- there are attributes of a species that aren't "traits" but are known to affect a species susceptibility to various threats

- these include:

  1.  habitat (vegetation association and moisture)
  2.  soil type (mainly soil textural terms) - under Phytophthora for now
  3.  underlying geologic substrate - under Phytophthora for now
  4.  population size (not merged in yet because too little data)
  5.  population count (not merged in yet because too little data)

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
# list of actual habitat terms for each species, from all sources
habitat <- all_traits$traits %>%
  filter(trait_name == "habitat") %>%
  select(taxon_name, habitat = value) |>
  group_by(taxon_name) %>%
  mutate(habitat = paste0(habitat, collapse = " ")) %>%
  ungroup() %>%
  distinct()

# extract habitat scores for each threat
C4_grasses_habitat <- extract_categorical_trait(all_traits, "habitat", habitat_scores %>% select(trait_name, trait_value, habitat = C4_grasses), "C4_grasses_habitat_effect_score")
goats_habitat <- extract_categorical_trait(all_traits, "habitat", habitat_scores %>% select(trait_name, trait_value, habitat = goats), "goat_habitat_effect_score")
myrtle_rust_habitat <- extract_categorical_trait(all_traits, "habitat", habitat_scores %>% select(trait_name, trait_value, habitat = myrtle_rust), "myrtle_rust_habitat_effect_score")
phytophthora_habitat <- extract_categorical_trait(all_traits, "habitat", habitat_scores %>% select(trait_name, trait_value, habitat = phytophthora), "phytophthora_habitat_effect_score")
pigs_habitat <- extract_categorical_trait(all_traits, "habitat", habitat_scores %>% select(trait_name, trait_value, habitat = pigs), "pig_habitat_effect_score")
deer_habitat <- extract_categorical_trait(all_traits, "habitat", habitat_scores %>% select(trait_name, trait_value, habitat = deer), "deer_habitat_effect_score")
hymenachne_habitat <- extract_categorical_trait(all_traits, "habitat", habitat_scores %>% select(trait_name, trait_value, habitat = hymenachne), "hymenachne_habitat_effect_score")

habitat_moisture <- extract_categorical_trait(all_traits, "habitat_moisture", habitat_moisture_scores, "habitat_moisture_effect_score")

# not yet being used, but should be for Phytophthora scoring - TODO
soil_type_scored <- extract_categorical_trait(all_traits, "soil_type", soil_type_scores, "soil_type")

# high pH soils (limestone, other calcareous soils) & saline soils are inhospitable to phytophthora. These are assigned a susceptibility = 0, all other soils susceptibility = 1
geologic_substrate_scored <- extract_categorical_trait(all_traits, "geologic_substrate", geologic_substrate_scores, "geologic_substrate")
```

### generate list species

- native species, per APC
- threatened species, per SPRAT website (accessed 2026-08-17)

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
native_species_list <- APCalign::native_anywhere_in_australia(species = resources$APC_accepted$canonical_name, resources = resources) |>
  filter(native_anywhere_in_aus == "native")
  
threatened_tmp <- read_csv("~/GitHub/threatened_species_traits/data_extras/threatened_flora.csv")

threatened <- APCalign::create_taxonomic_update_lookup(threatened_tmp$`Scientific Name`, resources = resources, full = T, fuzzy_matches = F) |>
  select(accepted_name, suggested_name, original_name, everything()) |>
  mutate(taxon_name = ifelse(is.na(accepted_name), original_name, accepted_name)) |>
  select(taxon_name, family, genus, taxonomic_dataset) |>
  mutate(threatened_status = "listed") |>
  filter(!is.na(family)) |> #remove 3 non-vascular plants (algae, liverworts: Vanvoorstia bennettiana, Lychnothamnus barbatus, Pseudocephalozia paludicola)
  filter(!taxon_name %in% c("Nitella parooensis", "Nereia lophocladia"))

native_species <- resources$APC %>%
  filter(taxonomic_status == "accepted") |>
  filter(taxon_rank %in% c("species", "subspecies", "variety", "form")) |>
  select(taxon_name = canonical_name, genus, family, taxon_distribution) |>
  filter(taxon_name %in% native_species_list$species) |>
  full_join(threatened)
```

# Threat independent effects

## Regeneration ability

### Assign susceptibilities to all taxa

#### - amalgamate data sources

```{r, echo=FALSE, message=FALSE, warning=FALSE}
regeneration_ability_effects <- tribble(
  ~trait,                  ~effect_score, ~notes,
  "life_history",            0.25,          "annuals, ephemerals = faster regeneration",
  "woodiness",            0.25,          "herbaceous = faster regeneration",
  "regeneration",            0.25,          "herbs = faster regeneration",
  "plant_height_scaled",            0.25,          "shorter plants reach maturity faster"
  #"seed_dry_mass_scaled",   -0.3,          "small seeds disperse further",
  #"dispersers",  -0.3,          "species with more effective dispersal syndromes more likely to have seed for regeneration",
  #"seedbank_location",  -0.2,          "species with local seedbanks more likely to have seed for regeneration",
  #"vegetative_reproduction_ability",  -0.2,          "species that can reproduce vegetatively more likely to persist in a location from belowground organs",
  #"reproductive_maturity_scaled", -0.6, "shorter time to maturity, regenerates faster",
  #"lifespan_scaled", -0.6, "shorter lived, regenerates faster",
  #"breeding_system_scaled", -0.5, "don't require outcrossing, regenerates faster",
  #"pollination_syndrome_scaled", -0.5, "more specific pollinators, requires bigger population, less overall disturbance"
)

# create dataframe with relevant traits, attributes
regeneration_ability_tmp <- native_species %>%
  left_join(habitat, by = "taxon_name") %>%
  left_join(life_history, by = "taxon_name") %>%
  left_join(woodiness, by = "taxon_name") %>%
  left_join(regeneration, by = "taxon_name") %>%
  left_join(plant_height, by = "taxon_name")
```

#### - calculate susceptibility scores

1.  calculate "susceptibility effect scores" for traits - taxonomy, habitat don't come in here
2.  sum traits susceptibilities
3.  scale total traits susceptibility
4.  assign susceptibility bins, confidence bins to each taxon
5.  this is not currently merged in with threat-specific susceptibility scores

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# calculate "susceptibility effect scores" for habitat, traits, taxonomy
regeneration_ability <- regeneration_ability_tmp %>%

   # calculate confidence in susceptibility score, based on how much data available, weighted
    # by trait effect size.
      calculate_prop_traits_scored_scaled(trait_table = regeneration_ability_effects) %>%

  # add {trait}_susceptibility for every active trait in threat_independent_effects (life
  # history, woodiness, herb/shrub/tree, plant height).
  add_trait_susceptibility(regeneration_ability_effects) %>%

  mutate(
    # sum trait susceptibility effect scores
      total_trait_effect_score = rowSums(across(c(life_history_susceptibility, woodiness_susceptibility, regeneration_susceptibility, plant_height_scaled_susceptibility)), na.rm = TRUE)
  ) %>%

  mutate(
    # rescale trait susceptibility
    # decide on range by looking at distribution of values
      trait_effect_score_scaled = sigmoidal_rescale(total_trait_effect_score, to = c(-0.1, 0.1)),

    # no habitat susceptibility to combine with here - total susceptibility is just the scaled trait susceptibility
      regeneration_ability_susceptibility = trait_effect_score_scaled
  )

regeneration_ability <- regeneration_ability |>
  select(regeneration_ability_susceptibility,
         taxon_name,
         threatened_status) |>
  # a handful of taxa (e.g. Dendrobium bigibbum) carry 2 rows here, which turns any later
  # left_join() into an unexpected many-to-many match and silently inflates row counts -
  # keep the first row per taxon
  distinct(taxon_name, .keep_all = TRUE)
```

#### - summary table of scorings

```{r, echo=FALSE, warning=FALSE, message=FALSE}
regeneration_ability_effects %>%
  arrange(effect_score, trait) %>% 
  kable(caption = "Population regeneration ability") %>% 
  kable_classic()
```

#### - susceptibility scores density plot

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=4}
df_long <- regeneration_ability %>%
  select(susceptibility = regeneration_ability_susceptibility,
         threatened_status) |>
  mutate(category = "trait susceptibility")

make_2_panel_susceptibility_density_plot(df_long, "threat-independent regeneration ability")

```

## Species precariousness

- this section captures how precarious a species' current situation already is, based on how little range it has left - independent of any trait-based resilience (covered above) or any specific threat
- unlike the threat-specific sections above, this does not use trait correlations to predict susceptibility for all Australian taxa - it scores each species directly from its own extent of occurrence (EOO), area of occupancy (AOO), and range size values, so it can only be calculated for taxa data for one of these attributes (predominantly already-listed/assessed taxa)
- smaller EOO/AOO/range size = more precarious (rarer, more geographically restricted) species

### Assign susceptibilities to all taxa

#### - amalgamate data sources

```{r, echo=FALSE, message=FALSE, warning=FALSE}
precariousness_effects <- tribble(
  ~trait,                                 ~effect_score, ~notes,
  "extent_of_occurrence_sigmoid_scaled",  -0.5,          "smaller extent of occurrence = more precarious",
  "area_of_occupancy_sigmoid_scaled",     -0.5,          "smaller area of occupancy = more precarious",
  "geographic_range_sigmoid_scaled",     -0.5,          "smaller range size = more precarious"
)

# create dataframe with relevant range-size attributes
# no habitat or taxonomic component - precariousness is scored directly from a species' own range size
precariousness_susceptibility_tmp <- native_species %>%
  left_join(extent_of_occurrence, by = "taxon_name") %>%
  left_join(area_of_occupancy, by = "taxon_name") %>%
  left_join(geographic_range, by = "taxon_name")
```

#### - calculate susceptibility scores

1.  calculate susceptibility effect scores directly from EOO, AOO - no habitat or taxonomy susceptibility involved
2.  sum EOO, AOO susceptibilities
3.  scale total susceptibility
4.  assign susceptibility bins, confidence bins to each taxon

```{r, echo=FALSE, message=FALSE, warning=FALSE}

precariousness <- precariousness_susceptibility_tmp %>%

  # calculate confidence in susceptibility score, based on how much data available (no habitat
  # term here), weighted by trait effect size.
    calculate_prop_traits_scored_scaled(trait_table = precariousness_effects) %>%

  mutate(
    # susceptibility_confidence_trait_count = prop_traits_scored_trait_count,
    susceptibility_confidence = prop_traits_scored
  ) %>%

  # add {trait}_susceptibility for every trait in precariousness_effects (extent of occurrence,
  # area of occupancy)
  add_trait_susceptibility(precariousness_effects) %>%

  mutate(
    # sum trait susceptibility effect scores
      total_trait_effect_score = rowSums(across(c(extent_of_occurrence_sigmoid_scaled_susceptibility, area_of_occupancy_sigmoid_scaled_susceptibility, geographic_range_sigmoid_scaled_susceptibility)), na.rm = TRUE)
  ) %>%

 # taxa with no data on range size need to be removed, so their susceptibility is `unknown` not 0 (= minimum susceptibility)
  filter(total_trait_effect_score != 0) %>%

  mutate(
    # rescale trait susceptibility
    # decide on range by looking at distribution of values
      precariousness_susceptibility = sigmoidal_rescale(total_trait_effect_score, to = c(0, 0.1))
      ) |>
  select(precariousness_susceptibility, taxon_name, threatened_status)
```

#### - summary table of scorings

```{r, echo=FALSE, warning=FALSE, message=FALSE}
precariousness_effects %>%
  arrange(effect_score, trait) %>%
  kable(caption = "Trait effect scores for species precariousness, based on remaining range size (EOO, AOO).") %>%
  kable_classic()
```

#### - susceptibility scores density plot

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=4}
df_long <- precariousness |>
  select(susceptibility = precariousness_susceptibility,
         threatened_status) |>
  mutate(category = "trait susceptibility")

make_2_panel_susceptibility_density_plot(df_long, "species precariousness")
```

# Phytophthora cinnamomi

### Are any traits good predictors?

- Check if trait correlations exist for taxa where susceptibility is already known.

#### - create dataframe

- create table with susceptibility, trait values for taxa with known *Phytophthora cinnamomi* susceptibility

#### - Phytophthora cinnamomi specific attrbitutes, scoring values

- Certain species are known to have higher or lower susceptibility to Phytophthora cinnamomi.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
phytophthora_cinnamomi_susceptibility_scores %>% 
  filter(!str_detect(trait_value, " ")) %>% 
  arrange(phytophthora_cinnamomi_susceptibility) %>% 
  rename(`categorical value` = trait_value, ordinal_value = phytophthora_cinnamomi_susceptibility)  %>% 
  kable(caption = "Phytophthora cinnamomi susceptibility values") %>% 
  kable_classic()
```

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
phytophthora_cinnamomi_susceptibility_data <- (all_traits %>% join_taxa(vars = "family"))$traits %>%
  filter(trait_name == "phytophthora_cinnamomi_susceptibility") %>%
  select(taxon_name, family, trait_value = value) %>%
  left_join(phytophthora_cinnamomi_susceptibility_scores, by = join_by(trait_value)) %>%
  group_by(taxon_name) %>%
  mutate(
    phytophthora_cinnamomi_susceptibility = mean(phytophthora_cinnamomi_susceptibility, na.rm = T)) %>% 
  ungroup() %>%
  distinct() %>% 
  left_join(life_history, by = "taxon_name") %>%
  left_join(woodiness, by = "taxon_name") %>%
  left_join(seed_dry_mass, by = "taxon_name") %>%
  left_join(leaf_mass_per_area, by = "taxon_name") %>%
  left_join(leaf_P_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_N_per_dry_mass, by = "taxon_name") %>%
  left_join(nitrogen_fixing, by = "taxon_name")%>%
  left_join(resprouting_capacity, by = "taxon_name") %>%
  left_join(storage_organ_fleshy, by = "taxon_name") %>%
  left_join(bud_bank_location_scaled, by = "taxon_name") 
```

#### - check for trait correlations

- identify strong susceptibility-trait correlations for taxa with known susceptibility

- Note:

  1.  Not considering traits that might matter but are entirely linked to taxonomy (i.e. types of roots/root associations)
  2.  Not considering traits that are known to matter, but are known for so few species to be useless as predictors (constitutive and induced responses to infection; concentration of secondary metabolites in plant)

- In the output:

  1.  a positive coefficient indicates a higher trait value in more susceptible taxa
  2.  for numeric traits, the magnitude of the coefficient indicates the change in response variable for a "1 unit" increase in susceptibility, where susceptibility is on a scale of 0--3.
  3.  for numeric traits, the magnitude of the coefficient must be intepreted in conjunction with knowing the scaling for the response variable - different for each trait; the susceptibility is on a scale of 0--3.
  4.  Also in this chart are correlations with soil type/texture and geologic substrate. As also indicated in the literature, it isn't that taxa growing of "resistant substrates/soils" are themselves resistant, but that Phytophthora doesn't survive in these locations. Therefore, this information can be used to assess the **RISK** of a species to Phytophthora, but not its **SUSCEPTIBILITY** to Phytophthora.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
correlations_with_possible_traits <- bind_rows(
  check_trait_correlations(
    phytophthora_cinnamomi_susceptibility_data, "phytophthora_cinnamomi_susceptibility",
    c("life_history", "nitrogen_fixing", "resprouting_capacity", "woodiness", "bud_bank_location_scaled", "storage_organ_fleshy"),
    r2_threshold = 0.03),
  check_trait_correlations(
    phytophthora_cinnamomi_susceptibility_data, "phytophthora_cinnamomi_susceptibility",
    c("seed_dry_mass", "leaf_mass_per_area", "leaf_N_per_dry_mass", "leaf_P_per_dry_mass"),
    r2_threshold = 0.03, use_log10 = TRUE)
) %>%
  arrange(-as.numeric(r2))

Phytophthora_kable_table <- correlations_with_possible_traits %>% kableExtra::kable(caption = "Correlation between Phytophthora susceptibililty and trait value, for species with known susceptibility")

kableExtra::kable_classic(Phytophthora_kable_table) %>%
  kableExtra::kable_styling(full_width = FALSE)

phytophthora_traits_to_use <- correlations_with_possible_traits %>% filter(p_value < 0.05, r2 > 0.03)
```

### How important is taxonomy? (vs traits...)

- Family matters more than traits
- Although some moderate correlations with traits exist, knowing a plant's family is a much better predictor of its susceptibility - at least visually

```{r, echo=FALSE, message=FALSE, warning=FALSE}

# total taxa currently accepted per family/genus (all of Australia's flora, not
# just this training dataset) - used below to identify "non-diverse" families/
# genera where even a small tested sample (n <= 5) already represents most/all
# of the taxa that exist in that group, so a small n there isn't the same susceptibility
# as a small n out of a large, diverse family/genus
family_totals <- resources$APC_accepted %>%
  count(family, name = "family_total_taxa")

genus_totals <- resources$APC_accepted %>%
  count(genus, name = "genus_total_taxa")

by_family <- phytophthora_cinnamomi_susceptibility_data %>%
  filter(!is.na(phytophthora_cinnamomi_susceptibility)) %>%
  group_by(family) %>%
  mutate(
    phytophthora_cinnamomi_susceptibility = mean(phytophthora_cinnamomi_susceptibility),
    taxa_scored_per_family = n()
    ) %>%
  ungroup() %>%
  distinct(family, taxa_scored_per_family, phytophthora_cinnamomi_susceptibility) %>%
  left_join(family_totals, by = "family") %>%
  arrange(-phytophthora_cinnamomi_susceptibility)

# For families Fabaceae and Myrtaceae there is enough within-family training-data
# to compare susceptibility differences at the genus-level
# And indeed there is enough genus-level variation to assign different susceptibility scores
# for certain genera - see below
by_genus <- phytophthora_cinnamomi_susceptibility_data %>%
  mutate(genus = stringr::word(taxon_name, start = 1, end = 1)) %>%
  filter(!is.na(phytophthora_cinnamomi_susceptibility)) %>%
  filter(family %in% c("Fabaceae", "Myrtaceae")) %>%
  group_by(genus) %>%
  mutate(
    phytophthora_cinnamomi_susceptibility = mean(phytophthora_cinnamomi_susceptibility),
    taxa_scored_per_genus = n()
    ) %>%
  ungroup() %>%
  distinct(genus, taxa_scored_per_genus, phytophthora_cinnamomi_susceptibility) %>%
  left_join(genus_totals, by = "genus") %>%
  arrange(-phytophthora_cinnamomi_susceptibility)

family_histograms <- phytophthora_cinnamomi_susceptibility_data %>%
  filter(family %in% by_family$family) %>%
  filter(!is.na(phytophthora_cinnamomi_susceptibility)) %>%
  left_join(by_family %>% select(family, taxa_scored_per_family)) %>%
  filter(taxa_scored_per_family > 5, !is.na(family)) %>%
  group_by(family, phytophthora_cinnamomi_susceptibility) %>%
  mutate(
    prop_taxa = n() / taxa_scored_per_family,
    count_taxa = n()
    ) %>%
  ungroup() %>%
  distinct(family, phytophthora_cinnamomi_susceptibility, prop_taxa, count_taxa, taxa_scored_per_family) %>%
  arrange(family, phytophthora_cinnamomi_susceptibility) %>%
  mutate(
    family2 = paste0(family, " (n=", taxa_scored_per_family,")")
  )

ggplot(data = family_histograms) +
  geom_col(aes(x = phytophthora_cinnamomi_susceptibility, y = prop_taxa), width = .9) +
  facet_wrap(~family2) +
  theme_minimal() +
  theme(
    axis.line.x.bottom = element_line(),
    axis.line.y.left = element_line()
  ) +
  ylab("prop. taxa tested (proportion within taxa tested within each family") +
  xlab("Phytophthora cinnamomi susceptibility score (higher = more susceptible)")
```

### Assign susceptibilities to all taxa

- first steps are based on taxonomy and habitat:
  1.  for taxa from well-sampled families, include a taxonomic adjustment factor *as a starting point, have the average susceptibility within the family, rescaled from 0--1.*
  2.  taxa that live on calcareous (limestone, basic, alkaline) soils are not susceptible
  3.  plants that live on clay soils are less susceptible - but this needs to be a down-weighted effect
- second steps will be to adjust these susceptibilities based on traits
  1.  need to decide on r\^2 threshold, below which the uncertainty is too great to include (possibly true for all traits where we have data!)
  2.  it would be great to have information on Proteaceae which explicitly do not form cluster roots or Ericaceae that do not form root hairs (or have very transient root hairs); also possible that cord roots increase susceptibility in Daviseia; but that doesn't seem to be available

NOTE: LMA emerges as having a significant correlation, but not including LMA effect score, because the effect is that very low LMA (thin leaved) taxa have lower susceptibility whereas no effect among higher LMA taxa. This is presumably about woodiness/life history, and doesn't make sense to also include effect of LMA

#### - amalgamate data sources

```{r, echo=FALSE, message=FALSE, warning=FALSE}
phytophthora_trait_effects <- tribble(
  ~trait,                  ~effect_score, ~notes,
  "woodiness",     0.3,         "herbaceous species at very low susceptibility; woody species at variable susceptibility",
  "resprouting_capacity",   -0.25,          "ability to resprout slows detrimental effects, especially if phytophthora  controlled at a site or site not always hospitable to phytophthora",
  "storage_organ_fleshy",  -0.2,          "species with a fleshy storage organ rarely susceptible; plants with woody storage organ or no storage organ quite variable in susceptibility",
  "leaf_P_per_dry_mass_scaled", -0.15,   "high leaf P = lower susceptibility, either because it is correlated with high soil P or high secondary metabolite content, but of which lower susceptibility",
  "leaf_N_per_dry_mass_scaled", -0.15,   "high leaf N = lower susceptibility, either because it is correlated with high soil N or high secondary metabolite content, but of which lower susceptibility",
  "reproductive_maturity_scaled",-0.3, "plants with shorter life cycles less impacted"
)

# McDougall (2024) explicitly lists species considered at susceptibility of extinction from Phytophthora cinnamomi; species listed here are explicitly assigned a susceptibility = 1 (the highest value), over-writing the trait/habitat modelled value and the known-susceptibility score.
phytophthora_extinction_risk <- read_csv("~/GitHub/threatened_species_traits/data_extras/phytophthora_cinnamomi_extinction_risk_McDougall_2024.csv") |>
  filter(`Current priority score` >= 80)

phytophthora_extinction_risk <- APCalign::create_taxonomic_update_lookup(phytophthora_extinction_risk$`Scientific Name`, resources = resources) |> 
  select(taxon_name = suggested_name) |>
  mutate(
    taxon_name = str_replace(taxon_name, " \\[.*", ""))


# for families and genera where species with known phytophthora susceptibility
# have consistently low or high susceptibilities, all species in the
# family/genus are assigned a low/high `taxonomic_effect_score`
# require > 5 tested taxa before trusting a family/genus average enough to
# apply it to every (also-untested) species in that group - the same
# threshold already used for the family_histograms plot above; without it, a
# single tested species (n = 1) could swing an entire family/genus to the
# scoring ceiling/floor
# EXCEPTION: small/non-diverse families or genera are retained even with <= 5
# tested taxa, provided that sample already covers >= 50% of all taxa
# currently accepted in that group (family_total_taxa/genus_total_taxa, from
# resources$APC_accepted) - there a small n is close to a full census of the
# group, not a risky handful drawn from many untested relatives
taxonomic_effect_scores_by_family <- by_family %>%
  filter(taxa_scored_per_family > 5 | taxa_scored_per_family / family_total_taxa >= 0.5) %>%
  filter(phytophthora_cinnamomi_susceptibility < 0.7 | phytophthora_cinnamomi_susceptibility >= 1.7) %>%
  select(family, phytophthora_cinnamomi_susceptibility, taxa_scored_per_family) %>%
  left_join(native_species %>% select(taxon_name, family), by = "family") %>%
  mutate(taxonomic_effect_score_notes = "family-level score applied")

taxonomic_effect_scores_by_genus <- by_genus %>%
  filter(taxa_scored_per_genus > 5 | taxa_scored_per_genus / genus_total_taxa >= 0.5) %>%
  filter(phytophthora_cinnamomi_susceptibility < 0.8 | phytophthora_cinnamomi_susceptibility >= 1.5) %>%
  select(genus, phytophthora_cinnamomi_susceptibility, taxa_scored_per_genus) %>%
  left_join(native_species %>% select(taxon_name, genus, family), by = "genus") %>%
  mutate(taxonomic_effect_score_notes = "genus-level score applied")

# score each taxon using the same thresholds as the filter it passed through above
# (family: 0.7/1.7, genus: 0.8/1.5) rather than one uniform pair for both, so every
# row that qualified for a taxon-level score above actually receives a non-zero one
taxonomic_effect_scores <- taxonomic_effect_scores_by_family %>%
  bind_rows(taxonomic_effect_scores_by_genus) %>%
  mutate(
    taxonomic_effect_score = case_when(
      taxonomic_effect_score_notes == "family-level score applied" & phytophthora_cinnamomi_susceptibility < 0.7 ~ -0.5*(1-phytophthora_cinnamomi_susceptibility),
      taxonomic_effect_score_notes == "family-level score applied" & phytophthora_cinnamomi_susceptibility >= 1.7 ~ 0.8*(phytophthora_cinnamomi_susceptibility/3),
      taxonomic_effect_score_notes == "genus-level score applied" & phytophthora_cinnamomi_susceptibility < 0.8 ~ -0.5*(1-phytophthora_cinnamomi_susceptibility),
      taxonomic_effect_score_notes == "genus-level score applied" & phytophthora_cinnamomi_susceptibility >= 1.5 ~ 0.8*(phytophthora_cinnamomi_susceptibility/3),
      TRUE ~ 0
    ),
    # unify the two n-tested columns (only one is ever non-NA per row, depending
    # on whether the family- or genus-level score applied) so downstream tables
    # can report how many taxa each taxonomic effect score is actually based on
    n_tested = coalesce(taxa_scored_per_family, taxa_scored_per_genus)
  )

# create dataframe with relevant traits, attributes
phytophthora_assigned_susceptibility <- native_species %>%
  left_join(phytophthora_cinnamomi_susceptibility %>% select(taxon_name, phytophthora_cinnamomi_susceptibility), by = "taxon_name") %>%
  left_join(taxonomic_effect_scores %>% select(taxonomic_effect_score, taxon_name), by = "taxon_name") %>%
  left_join(geologic_substrate_scored %>% rename(substrate_effect_score = geologic_substrate), by = "taxon_name") %>%
  left_join(soil_type_scored %>% rename(soil_texture_effect_score = soil_type), by = "taxon_name") %>%
  left_join(habitat, by = "taxon_name") %>%
  left_join(phytophthora_habitat, by = "taxon_name") %>%
  left_join(resprouting_capacity, by = "taxon_name") %>%
  left_join(reproductive_maturity, by = "taxon_name") %>%
  left_join(leaf_P_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_N_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_mass_per_area, by = "taxon_name") %>%
  left_join(woodiness, by = "taxon_name") %>%
  left_join(storage_organ_fleshy, by = "taxon_name") %>%
  # phytophthora_extinction_risk_flag is informational only (reported as its own column
  # in the export and in the report's Table 9) - it deliberately does NOT feed into
  # total_individual_susceptibility_overwrite or susceptibility_confidence below. It used
  # to override the modelled score to the dataset ceiling; Elizabeth + collaborator decided
  # extinction risk (McDougall 2024) shouldn't override the trait/taxonomy/known-susceptibility
  # model the way known susceptibility does - it's flagged for the reader instead.
  left_join(phytophthora_extinction_risk %>% mutate(phytophthora_extinction_risk_flag = TRUE),
            by = "taxon_name")
```

#### - calculate susceptibility scores

```{r, echo=FALSE, message=FALSE, warning=FALSE}
  phytophthora_bin_thresholds <- c(
      "5_extremely_high" = 0.4,
      "4_high"           = 0.25,
      "3_moderate"       = 0.15,
      "2_low"            = 0.1,
      "1_no_susceptibility"        = 0
    )

# calculate "susceptibility effect scores" for habitat, traits, taxonomy
phytophthora_assigned_susceptibility <- phytophthora_assigned_susceptibility %>%
  
    # calculate confidence in susceptibility score, based on how much data available, weighted
    # by trait effect size. (Comparison against the unweighted trait-count version
    # checked out - it shifted susceptibility_confidence by -0.08 to 0.2)
      calculate_prop_traits_scored_scaled(trait_table = phytophthora_trait_effects) %>%

  mutate(
    # default habitat_susceptibility is set to is 0.5
      habitat_susceptibility = ifelse(!is.na(phytophthora_habitat_effect_score), phytophthora_habitat_effect_score, .5),

    # species growing on calcareous/inhospitable geologic substrate scales habitat_susceptibility
    # down by 10x, regardless of the species' own habitat term or trait scores because the
    # pathogen can't survive there
    # this only applies when the substrate_effect_score is low,
    # to avoid overrides when the plant grows on many substrates, only one of them inhospitable
      habitat_susceptibility = ifelse(!is.na(substrate_effect_score) & substrate_effect_score < 0.35,
                            habitat_susceptibility * 0.1, habitat_susceptibility),

    # confidence drops to 0.15*prop_traits_scored (LOW) only if there's no habitat signal at
    # all - neither a direct habitat score nor a known inhospitable substrate
      susceptibility_confidence = ifelse(
        is.na(phytophthora_habitat_effect_score) &
          !(!is.na(substrate_effect_score) & substrate_effect_score < 0.35),
        0.15*prop_traits_scored, 0.6*prop_traits_scored),

    # confidence becomes 0.8 if known susceptibility
      susceptibility_confidence_overwrite = ifelse(!is.na(phytophthora_cinnamomi_susceptibility),
                               0.8, susceptibility_confidence),

    # Future possible addition:
    # while calcareous substrates are currently scored as reducing phytophtora habitat susceptibility,
    # soils high in organic matter or loam which are also inhospitable to phytophthora
    # should also have a smaller effect score; these data do not currently exist

        taxonomic_susceptibility = calculate_susceptibility("taxonomic_effect_score", taxonomic_effect_score, .)
  ) %>%

    # add {trait}_susceptibility for every trait in phytophthora_trait_effects (leaf P, leaf N,
    # woodiness, resprouting capacity, fleshy storage organ)
    add_trait_susceptibility(phytophthora_trait_effects) %>%

    mutate(
    # sum together trait and taxonomic effect scores. taxonomic_susceptibility is weighted 5x a
    # normal trait
      total_trait_effect_score = rowSums(across(c(
          leaf_P_per_dry_mass_scaled_susceptibility,
          leaf_N_per_dry_mass_scaled_susceptibility, resprouting_capacity_susceptibility,
          woodiness_susceptibility, storage_organ_fleshy_susceptibility
        )), na.rm = TRUE) + 5*coalesce(taxonomic_susceptibility, 0),

    # rescale trait susceptibility
    # decide on range by looking at distribution of values and deciding,
    # how much traits are "allowed" to offset habitat susceptibility
    # to ensure actual trait effect scores aren't too clumped,
    # often need a broader distribution than 0--1 and often need to recentre the distribution
      trait_effect_score_scaled = sigmoidal_rescale(total_trait_effect_score, to = c(0.05, 1.05)),

    # now add in known susceptibility - showing
    # multiple scaled trait effect score susceptibility by habitat susceptibility
      total_individual_susceptibility = habitat_susceptibility*trait_effect_score_scaled
  )
  
  phytophthora_assigned_susceptibility <- phytophthora_assigned_susceptibility %>%
    mutate(

    # There are four circumstances where a species' modeled `total_individual_susceptibility`
    # (from traits/habitat/taxonomy) is overwritten based on known susceptibility data:
    # 1. known susceptibility = 0 (resistant) -> damp the modelled score to 10% of itself, not
    #    replace it outright - habitat exposure can still nudge a confidently-resistant taxon up
    #    slightly, but nowhere near as far as the undamped model (which, for taxa with only 1-2
    #    traits scored, can land "extremely high" despite a strongly protective taxonomic score -
    #    see Acacia ataxiphylla subsp. magna, known = 1, taxonomic_effect_score = -0.32, but only
    #    woodiness scored, and habitat_susceptibility = 2 pushed the undamped model to "extremely high")
    # 2. known susceptibility = 1 (mild) -> damp the modelled score to 50% of itself, same logic
    #    as (1) but less severe, matching "mild" rather than "resistant"
    # 3. known susceptibility > 2 (severe, i.e. known = 3) -> double the modelled score
    # 4. known susceptibility (known/3) is HIGHER than the modelled score -> always overwrite
    #    upward, regardless of how high the known value itself is (this still applies to known
    #    values of 1.5-3; 0, 1 and >2 are now handled by rules 1-3 above instead of this rule,
    #    since known/3 was never higher than the modelled score for the taxa rules 1-3 target anyway)
    # The susceptibility_confidence is also increased to 0.8 rather than being based on the
    # trait-completeness-based value, as the susceptibility score is no longer trait-based.
    #
    # NOTE: the extinction-risk flag (McDougall 2024, `Current priority score` >= 80) used to
    # ALSO overwrite the modelled score (to the dataset ceiling, always) and force confidence to 1.
    # That's been deliberately removed (Elizabeth + collaborator, 2026-08-21): extinction risk is
    # now purely informational - reported as its own `phytophthora_extinction_risk_flag` column in
    # the export and as a column in the report's Table 9 - and no longer overrides the model the
    # way known susceptibility does.

      total_individual_susceptibility_overwrite = case_when(
        phytophthora_cinnamomi_susceptibility == 0 ~ total_individual_susceptibility * 0.1,
        phytophthora_cinnamomi_susceptibility == 1 ~ total_individual_susceptibility * 0.5,
        phytophthora_cinnamomi_susceptibility > 2.3 ~ total_individual_susceptibility * 2.5,
        !is.na(phytophthora_cinnamomi_susceptibility) & phytophthora_cinnamomi_susceptibility/3 > total_individual_susceptibility ~ phytophthora_cinnamomi_susceptibility/3,
        !is.na(phytophthora_cinnamomi_susceptibility) & phytophthora_cinnamomi_susceptibility/3 <= 0.2 ~ phytophthora_cinnamomi_susceptibility/3,
        TRUE ~ total_individual_susceptibility
      ),
      susceptibility_confidence = case_when(
        !is.na(phytophthora_cinnamomi_susceptibility) ~ 0.8,
        TRUE ~ susceptibility_confidence
      )
  ) %>%
    # merge in non-threat components
    left_join(precariousness |> select(-threatened_status)) |>
    left_join(regeneration_ability |> select(-threatened_status)) |>

    # for precariousness and regeneration, assign mean value for missing data
    mutate(
      precariousness_susceptibility = ifelse(is.na(precariousness_susceptibility),
                                   mean(precariousness_susceptibility, na.rm = TRUE),
                                   precariousness_susceptibility),
      
      regeneration_ability_susceptibility = ifelse(is.na(regeneration_ability_susceptibility),
                                   mean(regeneration_ability_susceptibility, na.rm = TRUE),
                                   regeneration_ability_susceptibility)
    )

  # rescale individual susceptibility to 0-1 across all taxa before combining with
  # regeneration/precariousness, so those two terms (fixed to a narrow +/-0.1-ish
  # range everywhere) have a consistent relative influence across every threat,
  # rather than one that depends on this threat's own raw susceptibility spread
  phytophthora_IS_range <- range(phytophthora_assigned_susceptibility$total_individual_susceptibility_overwrite, na.rm = TRUE)

  phytophthora_assigned_susceptibility <- phytophthora_assigned_susceptibility %>%
    mutate(
      total_individual_susceptibility_scaled = (total_individual_susceptibility_overwrite - phytophthora_IS_range[1]) / diff(phytophthora_IS_range)
    ) %>%

    # sum together 3 components, normalise
    mutate(
      population_susceptibility = (1*total_individual_susceptibility_scaled) + regeneration_ability_susceptibility + precariousness_susceptibility,
      population_susceptibility = ifelse(total_individual_susceptibility_overwrite == 0, 0, population_susceptibility)
    )

  # linearly rescale population_susceptibility to 0-1 across taxa
  phytophthora_susceptibility_range <- range(phytophthora_assigned_susceptibility$population_susceptibility, na.rm = TRUE)

  phytophthora_assigned_susceptibility <- phytophthora_assigned_susceptibility %>%
    mutate(
      population_susceptibility = (population_susceptibility - phytophthora_susceptibility_range[1]) / diff(phytophthora_susceptibility_range),

    # assign susceptibility bin (individual_susceptibility_bin removed - not reported to DCCEEW)
      population_susceptibility_bin = create_susceptibility_bins(population_susceptibility, bin_thresholds = phytophthora_bin_thresholds),
    ) |>

    # confidence-based plausible range around the numeric susceptibility score
    add_susceptibility_range(
      susceptibility_col = "population_susceptibility",
      confidence_col = "susceptibility_confidence",
      bin_thresholds = phytophthora_bin_thresholds
    )

  phytophthora_assigned_susceptibility_threatened <- phytophthora_assigned_susceptibility %>%
    filter(!is.na(threatened_status)) %>%
    select(taxon_name, genus, family, threatened_status,
           phytophthora_cinnamomi_susceptibility, phytophthora_extinction_risk_flag,
           total_individual_susceptibility, total_individual_susceptibility_overwrite,
           population_susceptibility, population_susceptibility_bin, susceptibility_confidence,
           population_susceptibility_min, population_susceptibility_max,
           population_susceptibility_bin_min, population_susceptibility_bin_max,
           habitat, habitat_susceptibility, trait_effect_score_scaled,
           taxonomic_susceptibility, prop_traits_scored,
           contains(c("susceptibility_effect_score")), everything()) %>%
    arrange(-population_susceptibility) |>
    write_csv("~/GitHub/threatened_species_traits/export/phytophthora_assigned_susceptibility_scores.csv", na = "")
```

#### - summary table of scorings

```{r, echo=FALSE, warning=FALSE, message=FALSE}
phytophthora_trait_effects %>%
    arrange(trait, effect_score) %>% 
    kable(caption = "Traits scored for Phytophthora cinnamomi. Note taxonomic effect scores are also applied as a proxy for traits that are closely tied to taxonomy, such as root structures and root types.") %>% 
    kable_classic()
  
family_component <- taxonomic_effect_scores %>%
    filter(!family %in% c("Fabaceae", "Myrtaceae")) %>%
    distinct(family, taxonomic_effect_score, phytophthora_cinnamomi_susceptibility, taxonomic_effect_score_notes, n_tested)

 taxonomic_effect_scores %>%
    filter(family %in% c("Fabaceae", "Myrtaceae")) %>%
    distinct(family, genus, taxonomic_effect_score, phytophthora_cinnamomi_susceptibility, taxonomic_effect_score_notes, n_tested) %>%
    bind_rows(family_component) %>%
    mutate(
      taxonomic_effect_score = round(taxonomic_effect_score, 2),
      phytophthora_cinnamomi_susceptibility = round(phytophthora_cinnamomi_susceptibility, 1)) %>%
    arrange(-taxonomic_effect_score, family, genus) %>%
    select(family, genus, n_tested, phytophthora_cinnamomi_susceptibility, taxonomic_effect_score, taxonomic_effect_score_notes) %>%
    kable(caption = "Families/genera assigned an explicit taxonomic effect score from known Phytophthora cinnamomi susceptibility (all others default to 0). n_tested is the number of taxa in that family/genus with a known susceptibility value informing the average; only groups with n_tested > 5 are scored, so a single unusual result can no longer swing an entire family/genus. Effect scores only applied if the family-average susceptibility score is < 0.7 or >= 1.7 (family level) / < 0.8 or >= 1.5 (genus level) on a 0-3 scale.") %>%
    kable_classic()
  
habitat_scores %>%
  select(habitat = trait_value, effect_score = phytophthora) %>%
  group_by(effect_score) %>%
  mutate(habitat = paste0(habitat, collapse = "; ")) %>%
  ungroup() %>%
  distinct() %>%
  arrange(effect_score) %>%
  kable(caption = "habitat scores for Phytophthora cinnamomi") %>% 
  kable_classic()

phytophthora_assigned_susceptibility_threatened |>
  count(population_susceptibility_bin) |>
  mutate(pct = round(100 * n / sum(n), 0)) |>
  arrange(desc(population_susceptibility_bin)) |>
  left_join(bin_thresholds_table(phytophthora_bin_thresholds), by = "population_susceptibility_bin") |>
  kable(caption = "taxa in each population susceptibility bin", col.names = c("Population susceptibility bin", "n", "%", "Threshold range")) %>%
  kable_classic()
```

#### - susceptibility scores density plot

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=13}
df_long_scaled <- phytophthora_assigned_susceptibility %>%
  select(`habitat susceptibility` = habitat_susceptibility,
         `trait and taxonomy susceptibility` = trait_effect_score_scaled,
         `combined individual susceptibility` = total_individual_susceptibility_overwrite,
         `mean population susceptibility` = population_susceptibility,
         `min population susceptibility` = population_susceptibility_min,
         `max population susceptibility` = population_susceptibility_max,
         threatened_status) %>%
  pivot_longer(cols = c(1:6),
               names_to = "category", values_to = "susceptibility")

make_8_panel_susceptibility_density_plot(df_long_scaled, paste0("Phytophthora cinnamomi"))
```

#### - save 4-panel report figure to PNG

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide', include=FALSE}
png("~/GitHub/threatened_species_traits/export/figures/phytophthora_4panel.png", width = 9, height = 6.5, units = "in", res = 150)
make_4_panel_susceptibility_density_plot(
  phytophthora_assigned_susceptibility, "Phytophthora cinnamomi",
  panel_cols = c("habitat_susceptibility", "trait_effect_score_scaled", "total_individual_susceptibility_overwrite", "population_susceptibility"),
  panel_labels = c("habitat susceptibility", "trait & taxonomy susceptibility", "individual susceptibility", "overall population susceptibility"),
  x_min = 0,
  x_max = 2
)
dev.off()
```

# Myrtle rust

### Are any traits good predictors?

#### - myrtle rust specific attrbitutes, scoring values

- Certain species are known to have higher or lower susceptibility to myrtle rust.
- Myrtle rust only impacts species growing in wet habitats.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
myrtle_rust_susceptibility_scores %>% 
  filter(!str_detect(trait_value, " ")) %>% 
  arrange(myrtle_rust_susceptibility) %>% 
  rename(`categorical value` = trait_value, ordinal_value = myrtle_rust_susceptibility)  %>% 
  kable(caption = "myrtle rust susceptibility values") %>% 
  kable_classic()
```

#### - create dataframe

- Create table with susceptibility, trait values for taxa with known susceptibility.

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
myrtle_rust_susceptibility_data <- myrtle_rust_susceptibility %>%
  left_join(dispersers, by = "taxon_name") %>%
  left_join(seed_dry_mass, by = "taxon_name") %>%
  left_join(leaf_mass_per_area, by = "taxon_name") %>%
  left_join(resprouting_capacity, by = "taxon_name")%>%
  left_join(reproductive_maturity, by = "taxon_name") %>%
  left_join(plant_height, by = "taxon_name") %>%
  left_join(leaf_length, by = "taxon_name") %>%
  left_join(leaf_hairs_adult_leaves, by = "taxon_name") %>%
  left_join(leaf_glaucousness, by = "taxon_name") %>%
  left_join(leaf_N_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_P_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_phenol_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_tannin_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_lignin_per_dry_mass, by = "taxon_name") %>%
  left_join(bud_bank_location_scaled, by = "taxon_name") %>%
  left_join(storage_organ_present, by = "taxon_name") %>%
  left_join(resources$APC %>% select(taxon_name = canonical_name, genus, family) |> distinct(taxon_name, .keep_all = T))
```

#### - check trait correlations

- For traits identified in the literature as being plausibly relevant, check if trait correlations exist for taxa with known susceptibility ("training data").

- Identify those traits for which there is evidence that they do affect susceptibility.

- Note:

  1.  Not considering traits that are known to matter, but are known for so few species to be useless as predictors (constitutive and induced responses to infection; concentration of secondary metabolites in plant); most of the data on plant metabolite content cannot be compared across studies because of different methodologies and therefore have not been included.
  2.  The traits `stomatal_distribution`, `leaf_glaucousness` and `leaf_hairiness` are very much expected to matter and this information should be available for sufficient taxa to asses whether there is a significant correlation with susceptibility, but currently these traits are scored for very few taxa.
  3.  There are traits including `leaf_wax_content`, `leaf_wax_structure` that might matter, but data exist for quite few taxa.

- In the output:

  1.  For myrtle rust susceptibility, a positive coefficient indicates a higher trait value in more susceptible taxa.
  2.  For numeric traits, the magnitude of the coefficient must be interpreted in conjunction with knowing the scaling for the response variable - different for each trait.
  3.  For categorical traits, each trait has different ordinal scores, sometimes on a scale of 0--1, but up to 0--5; see tables above.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
correlations_with_possible_traits_myrtle_rust <- bind_rows(
  check_trait_correlations(
    myrtle_rust_susceptibility_data, "myrtle_rust_susceptibility",
    c("resprouting_capacity", "leaf_glaucousness", "leaf_hairs_adult_leaves", "bud_bank_location_scaled", "storage_organ_present"),
    r2_threshold = 0.03),
  check_trait_correlations(
    myrtle_rust_susceptibility_data, "myrtle_rust_susceptibility",
    c("seed_dry_mass", "leaf_mass_per_area", "plant_height", "leaf_length", "leaf_N_per_dry_mass", "leaf_P_per_dry_mass", "leaf_lignin_per_dry_mass", "leaf_phenol_per_dry_mass", "leaf_tannin_per_dry_mass"),
    r2_threshold = 0.04, use_log10 = TRUE)
)

myrtlerust_kable_table <- correlations_with_possible_traits_myrtle_rust %>%
  arrange(-as.numeric(r2)) %>%
  kableExtra::kable(caption = "Correlation between myrtle rust susceptibililty and trait value, for species with known susceptibility")

kableExtra::kable_classic(myrtlerust_kable_table) %>%
  kableExtra::kable_styling(full_width = FALSE)

```

### How important is taxonomy? (vs traits...)

considering whether genus matters as much as - or more than traits

- not sufficiently clear to use this, although there are some genera where all species tested are susceptible and others where there are some resistant species and many partially resistant species

```{r, echo=FALSE, warning=FALSE, message=FALSE}
by_genus <- myrtle_rust_susceptibility_data %>%
  filter(!is.na(myrtle_rust_susceptibility)) %>%
  group_by(genus) %>%
  mutate(
    myrtle_rust_susceptibility = mean(myrtle_rust_susceptibility),
    taxa_scored_per_genus = n()
    ) %>%
  ungroup() %>%
  distinct(genus, taxa_scored_per_genus, myrtle_rust_susceptibility) %>%
  filter(taxa_scored_per_genus > 3, !is.na(genus)) %>%
  arrange(-myrtle_rust_susceptibility)

genus_histograms <- myrtle_rust_susceptibility_data %>%
  filter(genus %in% by_genus$genus) %>%
  filter(!is.na(myrtle_rust_susceptibility)) %>%
  left_join(by_genus %>% select(genus, taxa_scored_per_genus)) %>%
  group_by(genus, myrtle_rust_susceptibility) %>%
  mutate(
    prop_taxa = n() / taxa_scored_per_genus,
    count_taxa = n()
    ) %>%
  ungroup() %>%
  distinct(genus, myrtle_rust_susceptibility, prop_taxa, count_taxa, taxa_scored_per_genus) %>%
  arrange(genus, myrtle_rust_susceptibility) %>%
  mutate(
    genus2 = paste0(genus, " (n=", taxa_scored_per_genus,")")
  )

ggplot(data = genus_histograms) +
  geom_col(aes(x = myrtle_rust_susceptibility, y = prop_taxa), width = .9) +
  facet_wrap(~genus2) +
  theme_minimal() +
  theme(
    axis.line.x.bottom = element_line(),
    axis.line.y.left = element_line()
  ) +
  ylab("prop. of taxa tested") +
  xlab("myrtle rust susceptibility score (higher = more susceptible)")
```

### Assign susceptibilities to all taxa

- first steps are based on taxonomy and habitat:

  1.  only Myrtaceae are susceptible
  2.  only taxa that live in "wet" environments are susceptible

```         
*ISSUE*: How to score taxa where habitat information isn't known; currently have data for \~50% of Myrtaceae. Right now we are taking the conservative approach and scoring all those species as being susceptible. Habitat is the first trait we plan to gap fill, but we might not end up with data for all taxa.
```

- second steps will be to adjust these binary susceptibilities based on traits

  1.  need to decide on r2 threshold, below which the uncertainty is too great to include (possibly true for all traits where we have data!)
  2.  there are additional traits which might well be important, but need to bolster their coverage in database to be certain: **leaf glaucousness**, **leaf wax content**, **leaf hairiness**

#### - amalgamate data sources

```{r, echo=FALSE, message=FALSE, warning=FALSE}
myrtle_rust_trait_effects <- tribble(
  ~trait,                  ~effect_score, ~notes,
  "leaf_length_scaled",     -0.2,         "not in literature, but no small leaf resistant species; this is a trait that is mostly age-independent, so if a good predictor, particularly useful",
  "plant_height_scaled",   -0.4,          "taller plants = lower impact, because taller plant canopies further from moist ground",
  "resprouting_capacity",  -0.3,          "resprouters = slower impact",
  "bud_bank_location_scaled", -0.3,   "bud bank above the ground surface improves chance of resprouting vegetation surviving",
  "leaf_lignin_per_dry_mass_scaled", -0.2, "higher leaf lignin = lower impact" # was "leaf_lignin_per_dry_mass" (missing _scaled) - that pointed at the raw, character-typed column and never actually matched an effect score lookup either, so this trait has silently never contributed to myrtle rust susceptibility until now; fixed at the user's request
)

# read in species list of epiphytic orchids with known preference for Myrtaceae species as hosts (Zimmer 2023)
# these orchids have a `taxonomic_susceptibility` applied to them, as they will
# be impacted if their host dies from myrtle rust
orchids_at_susceptibility <- read_csv("~/GitHub/threatened_species_traits/data/Zimmer_2023/data.csv") %>%
  mutate(
    taxon_name = stringr::str_extract(`Orchid species`, "[^\\(]+"),
    taxon_name = stringr::str_replace(taxon_name, "\\*",""),
    taxon_name = stringr::str_trim(taxon_name)) %>%
  select(taxon_name, score) %>%
  mutate(orchid_susceptibility = case_when(
    score == "high" ~ 1,
    score == "moderate" ~ 0.6,
    score == "low" ~ 0.3
  ))

orchids_summary <- orchids_at_susceptibility %>%
  select(taxon = taxon_name, effect_score = orchid_susceptibility) %>%
  group_by(effect_score) %>%
  mutate(taxon = paste0(taxon, collapse = ", ")) %>%
  ungroup() %>%
  distinct() %>%
  mutate(family = "Orchidaceae", taxon_rank = "species")

myrtle_rust_taxonomic_effects <- 
  tibble(
    taxon = "Myrtaceae",
    taxon_rank = "family",
    effect_score = 1,
    family = "Myrtaceae"
    ) %>%
  bind_rows(orchids_summary) %>%
  bind_rows(
  tibble(
    taxon = "all other taxa",
    taxon_rank = NA,
    effect_score = 0.2,
    family = NA
  )
)

# create dataframe with relevant traits, attributes
myrtle_rust_assigned_susceptibility <- native_species %>%
  left_join(myrtle_rust_susceptibility, by = "taxon_name") %>%
  left_join(habitat, by = "taxon_name") %>%
  left_join(myrtle_rust_habitat, by = "taxon_name") %>%
  left_join(habitat_moisture, by = "taxon_name") %>%
  left_join(plant_height |> distinct(), by = "taxon_name") %>%
  left_join(leaf_length |> distinct(), by = "taxon_name") %>%
  left_join(reproductive_maturity, by = "taxon_name") %>%
  left_join(resprouting_capacity, by = "taxon_name") %>%
  left_join(bud_bank_location_scaled, by = "taxon_name") %>%
  left_join(leaf_lignin_per_dry_mass, by = "taxon_name") %>%
  left_join(orchids_at_susceptibility, by = "taxon_name")
```

#### - calculate susceptibility scores

```{r, echo=FALSE, message=FALSE, warning=FALSE}
 myrtle_rust_bin_thresholds <- c(
      "5_extremely_high" = 0.25,
      "4_high"           = 0.15,
      "3_moderate"       = 0.1,
      "2_low"            = 0.05,
      "1_no_susceptibility"        = 0
    )

# calculate "susceptibility effect scores" for habitat, traits, taxonomy
myrtle_rust_assigned_susceptibility <- myrtle_rust_assigned_susceptibility %>%
    # calculate confidence in susceptibility score, based on how much data available, weighted
    # by trait effect size. 
      calculate_prop_traits_scored_scaled(trait_table = myrtle_rust_trait_effects) %>%
  mutate(
    # starting susceptibility for Myrtaceae is 1; for all other taxa = 0.2,
    # since other rainforest plants might be impacted by changes in community structure
    # also incorporate list of epiphytic orchids known to be at susceptibility
    # because they sometimes or always grow on Myrtaceae species
      taxonomic_susceptibility = case_when(
          family == "Myrtaceae" ~ 1,
          !is.na(orchid_susceptibility) ~ orchid_susceptibility,
          TRUE ~ 0.2),

    # there are taxa where we have a `habitat_moisture` score,
    # as well as, or instead of, a `habitat` score
    # If both exist, the "higher" score takes precedence
    # If only `habitat_moisture` exists, it is used over having a black `habitat_susceptibility`
      habitat_susceptibility = ifelse(
        is.na(myrtle_rust_habitat_effect_score) & !is.na(habitat_moisture_effect_score),
        habitat_moisture_effect_score,
        # if habitat moisture value higher, use it over habitat value
          ifelse(
            !is.na(myrtle_rust_habitat_effect_score) & !is.na(habitat_moisture_effect_score) &
              habitat_moisture_effect_score > myrtle_rust_habitat_effect_score,
            habitat_moisture_effect_score, myrtle_rust_habitat_effect_score)),

    # confidence drops to 0.15*prop_traits_scored (LOW) only if there's no data for `habitat` nor `habitat_moisture`  
      susceptibility_confidence = ifelse(is.na(habitat_susceptibility),
                               0.15*prop_traits_scored, 0.6*prop_traits_scored),

    # confidence becomes 0.8 if known susceptibility
      susceptibility_confidence_overwrite = ifelse(!is.na(myrtle_rust_susceptibility),
                               0.8, susceptibility_confidence),

    # merge habitat and taxonomy susceptibilities; a "0" in either means no susceptibility
      habitat_taxonomy_susceptibility = ifelse(is.na(habitat_susceptibility),
                                     taxonomic_susceptibility, taxonomic_susceptibility*habitat_susceptibility)
  ) %>%

    # the calculated trait effect scores only apply to Myrtaceae, not other rainforest plants
    # these plants are affected by myrtle rust through undefined the changes in community structure
    # (for instance, decrease in shade, increasing temperature)
    # not as a result of their traits
    # therefore, non-Myrtaceae taxa (including the at-risk orchids) get their
    # susceptibility via `habitat_taxonomy_susceptibility`  only
    add_trait_susceptibility(myrtle_rust_trait_effects, condition = .$family == "Myrtaceae") %>%

    mutate(
    # sum trait susceptibility effect scores
      total_trait_effect_score =  rowSums(across(c(plant_height_scaled_susceptibility, leaf_length_scaled_susceptibility, resprouting_capacity_susceptibility, bud_bank_location_scaled_susceptibility, leaf_lignin_per_dry_mass_scaled_susceptibility)), na.rm = TRUE),

    # rescale trait susceptibility
    # decide on range by looking at distribution of values and deciding, 
    # how much traits are "allowed" to offset habitat susceptibility
      trait_effect_score_scaled = sigmoidal_rescale(total_trait_effect_score, to = c(0.2, 1.15)), #0.4, 1.35

    # total susceptibility is the product of scaled trait susceptibility and habitat/taxonomy susceptibility
      total_individual_susceptibility = habitat_taxonomy_susceptibility * trait_effect_score_scaled,

    # for species with known susceptibility to myrtle rust from inoculation experiments,
    # the trait-modeled score is overwritten based on that known value:
    # - known = "high" (5, the top of the 0-5 known scale) -> boost to 1.5x the modelled score
    # - known = "medium"/"susceptible" (3) -> boost to 1.25x the modelled score
    # - known = "resistant"/"low" (0/1) and the modelled score is already < 1 -> damp down to
    #   known/3, regardless of trait-modeled score
      total_individual_susceptibility_overwrite = case_when(
        myrtle_rust_susceptibility == 5 ~ total_individual_susceptibility * 1.5,
        myrtle_rust_susceptibility == 3 ~ total_individual_susceptibility * 1.25,
        !is.na(myrtle_rust_susceptibility) & total_individual_susceptibility < 1 ~ myrtle_rust_susceptibility/3,
        TRUE ~ total_individual_susceptibility
      ),
    
    # if actual susceptibility is know, increase confidence
      susceptibility_confidence = ifelse(
        !is.na(myrtle_rust_susceptibility), 
        0.8, susceptibility_confidence
      )
    ) |>

    # merge in non-threat components
    left_join(precariousness |> select(-threatened_status)) |>
    left_join(regeneration_ability |> select(-threatened_status)) |>

    # for precariousness and regeneration, assign mean value for missing data
    mutate(
      precariousness_susceptibility = ifelse(is.na(precariousness_susceptibility),
                                   mean(precariousness_susceptibility, na.rm = TRUE),
                                   precariousness_susceptibility),
      
      regeneration_ability_susceptibility = ifelse(is.na(regeneration_ability_susceptibility),
                                   mean(regeneration_ability_susceptibility, na.rm = TRUE),
                                   regeneration_ability_susceptibility)
    )

  # rescale individual susceptibility to 0-1 across all taxa before combining with
  # regeneration/precariousness, so those two terms (fixed to a narrow +/-0.1-ish
  # range everywhere) have a consistent relative influence across every threat,
  # rather than one that depends on this threat's own raw susceptibility spread
  myrtle_rust_IS_range <- range(myrtle_rust_assigned_susceptibility$total_individual_susceptibility_overwrite, na.rm = TRUE)

  myrtle_rust_assigned_susceptibility <- myrtle_rust_assigned_susceptibility %>%
    mutate(
      total_individual_susceptibility_scaled = (total_individual_susceptibility_overwrite - myrtle_rust_IS_range[1]) / diff(myrtle_rust_IS_range)
    ) %>%

    # sum together 3 components, normalise
    mutate(
      population_susceptibility = (1*total_individual_susceptibility_scaled) + regeneration_ability_susceptibility + precariousness_susceptibility,
      #population_susceptibility = sigmoidal_rescale(total_trait_effect_score, to = c(0.0, 1)),
      population_susceptibility = ifelse(total_individual_susceptibility_overwrite == 0, 0, population_susceptibility)
    )

  # linearly rescale population_susceptibility to 0-1 across taxa
  myrtle_rust_susceptibility_range <- range(myrtle_rust_assigned_susceptibility$population_susceptibility, na.rm = TRUE)

  myrtle_rust_assigned_susceptibility <- myrtle_rust_assigned_susceptibility |>
    mutate(
      population_susceptibility = (population_susceptibility - myrtle_rust_susceptibility_range[1]) / diff(myrtle_rust_susceptibility_range),

    # assign susceptibility bin (individual_susceptibility_bin removed - not reported to DCCEEW)
      population_susceptibility_bin = create_susceptibility_bins(population_susceptibility, bin_thresholds = myrtle_rust_bin_thresholds)

    ) |>

    # confidence-based plausible range around the numeric susceptibility score
      add_susceptibility_range(
        susceptibility_col = "population_susceptibility",
        confidence_col = "susceptibility_confidence",
        bin_thresholds = myrtle_rust_bin_thresholds
      )

  myrtle_rust_assigned_susceptibility_threatened <- myrtle_rust_assigned_susceptibility |>
    filter(!is.na(threatened_status)) |>
    select(taxon_name, genus, family, threatened_status,
           myrtle_rust_susceptibility,
           total_individual_susceptibility, total_individual_susceptibility_overwrite,
           population_susceptibility, population_susceptibility_bin, susceptibility_confidence,
           population_susceptibility_min, population_susceptibility_max,
           population_susceptibility_bin_min, population_susceptibility_bin_max,
           habitat, habitat_susceptibility, trait_effect_score_scaled,
           taxonomic_susceptibility, prop_traits_scored,
           contains(c("susceptibility_effect_score")), everything()) |>
    arrange(-population_susceptibility) |>
    write_csv("~/GitHub/threatened_species_traits/export/myrtle_rust_susceptibility_scores.csv", na = "")
```

#### - summary table of scorings

```{r, echo=FALSE, warning=FALSE, message=FALSE}
 myrtle_rust_trait_effects %>%
    arrange(effect_score, trait) %>% 
    kable(caption = "Traits scored for myrtle rust.") %>% 
    kable_classic()

myrtle_rust_taxonomic_effects %>%
    select(taxon, family, taxon_rank, effect_score) %>%
    arrange(-effect_score, family, taxon) %>%
    kable(caption = "Taxonomic effect scores for myrtle rust.") %>%
    kable_classic()

habitat_scores %>%
  select(habitat = trait_value, effect_score = myrtle_rust) %>%
  group_by(effect_score) %>%
  mutate(habitat = paste0(habitat, collapse = "; ")) %>%
  ungroup() %>%
  distinct() %>% 
  select(habitat, effect_score) %>%
  arrange(effect_score) %>%
  kable(caption = "Habitat effect scores for myrtle rust.") %>% 
  kable_classic()

habitat_moisture_scores %>%
  select(effect_score = habitat_moisture, habitat_moisture = trait_value) %>%
  group_by(effect_score) %>%
  mutate(habitat_moisture = paste0(habitat_moisture, collapse = "; ")) %>%
  ungroup() %>%
  distinct() %>% 
  select(habitat_moisture, effect_score) %>%
  arrange(effect_score) %>%
  kable(caption = "Habitat moisture effect scores for myrtle rust.") %>% 
  kable_classic()

myrtle_rust_myrtaceae_bins <- myrtle_rust_assigned_susceptibility_threatened |>
  filter(family == "Myrtaceae") |>
  count(population_susceptibility_bin, name = "n_myrtaceae") |>
  mutate(pct_myrtaceae = round(100 * n_myrtaceae / sum(n_myrtaceae), 1))

myrtle_rust_assigned_susceptibility_threatened |>
  count(population_susceptibility_bin) |>
  mutate(pct = round(100 * n / sum(n), 0)) |>
  arrange(desc(population_susceptibility_bin)) |>
  left_join(bin_thresholds_table(myrtle_rust_bin_thresholds), by = "population_susceptibility_bin") |>
  left_join(myrtle_rust_myrtaceae_bins, by = "population_susceptibility_bin") |>
  mutate(
    n_myrtaceae = tidyr::replace_na(n_myrtaceae, 0),
    pct_myrtaceae = tidyr::replace_na(pct_myrtaceae, 0)
  ) |>
  kable(caption = "taxa in each population susceptibility bin, and the subset within family Myrtaceae", col.names = c("Population susceptibility bin", "n", "%", "Threshold range", "n (Myrtaceae)", "% (Myrtaceae)")) %>%
  kable_classic()
```

#### - susceptibility scores density plot (Myrtaceae only)

Trait scoring only applies to Myrtaceae (see `add_trait_susceptibility(condition = family == "Myrtaceae")` above), so the other 87% of species all get `total_trait_effect_score == 0` `sigmoidal_rescale()` values are also computed from Myrtaceae only.

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=13}
df_long_scaled <- myrtle_rust_assigned_susceptibility %>%
  filter(family == "Myrtaceae") %>%
  select(`habitat susceptibility` = habitat_taxonomy_susceptibility,
         `trait and taxonomy susceptibility` = trait_effect_score_scaled,
         `combined individual susceptibility` = total_individual_susceptibility_overwrite,
         `mean population susceptibility` = population_susceptibility,
         `min population susceptibility` = population_susceptibility_min,
         `max population susceptibility` = population_susceptibility_max,
         threatened_status) %>%
  pivot_longer(cols = c(1:6),
               names_to = "category", values_to = "susceptibility")

make_8_panel_susceptibility_density_plot(df_long_scaled, paste0("myrtle rust, Myrtaceae only (sigmoidal_rescale, k = ", sigmoidal_rescale_k, ")"))
```

#### - save 4-panel report figure to PNG

Habitat & taxonomy susceptibility, individual susceptibility and population susceptibility are shown for all currently-listed taxa; trait susceptibility is masked to NA for non-Myrtaceae taxa before plotting (rather than filtering the whole dataframe to Myrtaceae, as the row above does) - trait scoring only applies to Myrtaceae, so the other 87% of listed taxa have no real trait-susceptibility value to show, but they still belong in the other three panels.

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide', include=FALSE}
myrtle_rust_plot_data <- myrtle_rust_assigned_susceptibility %>%
  filter(family == "Myrtaceae")

png("~/GitHub/threatened_species_traits/export/figures/myrtle_rust_4panel.png", width = 9, height = 6.5, units = "in", res = 150)
make_4_panel_susceptibility_density_plot(
  myrtle_rust_plot_data, "myrtle rust",
  panel_cols = c("habitat_taxonomy_susceptibility", "trait_effect_score_scaled", "total_individual_susceptibility_overwrite", "population_susceptibility"),
  panel_labels = c("habitat susceptibility", "trait susceptibility (Myrtaceae only)", "individual susceptibility", "overall population susceptibility"),
  x_min = 0,
  x_max = 1.6
)
dev.off()
```

# Goats

### Are any traits good predictors?

- Check if trait correlations exist for taxa where palatability/forage preference/toxicity is known.
- For goats, there are no lists of species that goats avoid -except linked already to specific traits.
- For instance, there is a list of 360 species growing in arid Australia, indicating their toxicity to goats; same species have palatibility values.
- Goats are known to eat "spiny acacias" (multiple references to this), but presumably the most spine-protected plants are less susceptible?
- Trees are resistant as adults, especially if they have thick bark (little data), but not as seedlings.
- Herbs much less impacted than shrubs, grasses.
- Succulence

#### - goat specific attributes, scoring values

- Certain species are known to be toxic to goats
- Certain species are known to be the most palatable to goats
- Goats prefer arid environments, especially rocky areas

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# print score tables
plant_goat_toxicity_scores %>% 
  filter(!str_detect(trait_value, " ")) %>% 
  arrange(plant_goat_toxicity) %>% 
  rename(`categorical value` = trait_value, ordinal_value = plant_goat_toxicity)  %>% 
  kable(caption = "plant goat toxicity values") %>% 
  kable_classic()

plant_goat_palatability_scores %>% 
  filter(!str_detect(trait_value, " ")) %>% 
  arrange(plant_goat_palatability) %>% 
  rename(`categorical value` = trait_value, ordinal_value = plant_goat_palatability)  %>% 
  kable(caption = "plant goat palatability values") %>% 
  kable_classic()
```

#### - create dataframe

```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
goat_susceptibility <- plant_goat_palatability %>%
  select(-genus) %>%
  left_join(resources$APC %>% select(family, genus, taxon_name = canonical_name) %>% distinct(), by = "taxon_name") %>%
  group_by(taxon_name) %>%
  mutate(
    plant_goat_palatability  = mean(plant_goat_palatability , na.rm = T)) %>% 
  ungroup() %>%
  distinct() %>% 
  left_join(leaf_water_content_per_area, by = "taxon_name") %>%
  left_join(life_history, by = "taxon_name") %>%
  left_join(woodiness, by = "taxon_name") %>%
  left_join(forb_graminoid, by = "taxon_name") %>%
  left_join(herb_shrub_tree, by = "taxon_name") %>%
  left_join(leaf_hairs_adult_leaves, by = "taxon_name") %>%
  left_join(leaf_mass_per_area, by = "taxon_name") %>%
  left_join(resprouting_capacity, by = "taxon_name") %>%
  left_join(leaf_P_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_N_per_dry_mass, by = "taxon_name") %>%
  left_join(plant_height, by = "taxon_name") %>%
  left_join(plant_spinescence, by = "taxon_name") %>%
  left_join(plant_physical_defence_structures, by = "taxon_name") %>%
  left_join(leaf_lignin_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_phenol_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_tannin_per_dry_mass, by = "taxon_name") %>%
  left_join(digestible_mass_per_dry_mass, by = "taxon_name") %>%
  left_join(leaf_crude_protein_per_dry_mass, by = "taxon_name") %>%
  left_join(metabolisable_energy, by = "taxon_name") %>%
  left_join(leaf_work_to_shear_adjusted, by = "taxon_name")
```

#### - check for trait correlations

```         
1.  Leaf tannins is coming out with higher tannins having higher palatability - the opposite of what it should be, becaise n = 5 (p-value not significant)
2.  Nitrogen fixing comes out as a strong negative correlation because all members of the genus Swainsona mapped in as toxic - and all are N-fixers. (This trait has been removed)
3.  Succulence (`leaf_water_content_per_area` ) should be positively correlated with palatability, but only scored for 4 species and happens to come out as a strong negative correlation (p-value not significant)
4.  `plant_spinescence` has the opposite correlation to what you would expect.
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
correlations_with_possible_traits_goats <- bind_rows(
  check_trait_correlations(
    goat_susceptibility, "plant_goat_palatability",
    c("life_history", "resprouting_capacity", "leaf_hairs_adult_leaves", "plant_spinescence", "plant_physical_defence_structures", "herb_shrub_tree", "forb_graminoid", "woodiness"),
    r2_threshold = 0.01),
  # other traits with too small sample sizes to bother with: c("leaf_phenol_per_dry_mass, "leaf_lignin_per_dry_mass", "leaf_tannin_per_dry_mass")
  check_trait_correlations(
    goat_susceptibility, "plant_goat_palatability",
    c("leaf_mass_per_area", "plant_height", "leaf_N_per_dry_mass", "leaf_P_per_dry_mass", "digestible_mass_per_dry_mass", "leaf_crude_protein_per_dry_mass", "metabolisable_energy", "leaf_work_to_shear_adjusted"),
    r2_threshold = 0.02, use_log10 = TRUE)
) %>%
  mutate(r2 = as.numeric(r2), coeff = as.numeric(coeff)) %>%
  arrange(-r2) %>%
  filter(!is.na(trait))

correlations_with_possible_traits_goats %>% 
  kableExtra::kable(caption = "Correlation between goat palatability and trait value, for species with known susceptibility. Positive coefficient indicates higher numeric/ordinal value = higher palatability") %>%
  kableExtra::kable_classic() %>%
  kableExtra::kable_styling(full_width = FALSE)
```

### Assign susceptibilities to all taxa

#### - amalgamate data sources

```{r, echo=FALSE, message=FALSE, warning=FALSE}
goat_trait_effects <- tribble(
  ~trait, ~effect_score,       ~notes,
  "woodiness",0.3,             "woody species more impacted",
  "life_history",0.4,             "perennial species more impacted",
  "forb",-0.15, "forbs are less impacted than graminoids",
  "resprouting_capacity", -0.25, "resprouting species are more likely to regenerate, persist; however, slight effect of palatable species more likely to be fire-killed",
  "plant_height_4m_peak_scaled", -0.1,  "worst height is 4 m, with decreasing impact above and below that",
  "bud_bank_location_scaled",-0.2,    "having a higher bud bank means new shoots more likely to persist",
  "bark_thickness_scaled",-0.2,   "higher bark thickness = lower impact",
  "reproductive_maturity_scaled",-0.3, "plants with shorter life cycles less impacted"
)

# taxonomic effect scores assigned for families with known high or low palatibility
goat_taxonomic_effects <- tribble(
  ~family, ~effect_score,
  "Chenopodiaceae",-0.1,
  "Restionaceae",0.1,
  "Poaceae", 0.1,
  "Cyperaceae", 0.1,
  "Juncaginaceae",0.1,
  "Juncaceae",0.1,
)

graminoids <- c("Restionaceae", "Poaceae", "Cyperaceae", "Juncaginaceae", "Juncaceae")

# create dataframe with relevant traits, attributes
goats_assigned_susceptibility <- native_species %>%
  left_join(habitat, by = "taxon_name") %>%
  left_join(goats_habitat, by = "taxon_name") %>%
  left_join(plant_height |> distinct(), by = "taxon_name") %>%
  left_join(resprouting_capacity, by = "taxon_name") %>%
  left_join(reproductive_maturity, by = "taxon_name") %>%
  left_join(bud_bank_location_scaled, by = "taxon_name") %>%
  left_join(life_history, by = "taxon_name") %>%
  left_join(bark_thickness, by = "taxon_name") %>%
  left_join(woodiness, by = "taxon_name") %>%
  left_join(plant_goat_palatability %>% select(-genus), by = "taxon_name") %>%
# graminoids (herbs within one of the graminoid families) are more impacted by goats than forbs (non-graminoid herbs)
  mutate(
    forb = ifelse(woodiness == 0 & family %in% graminoids, 0, NA),
    forb = ifelse(woodiness == 0 & !family %in% graminoids, 1, forb),
    taxonomic_susceptibility = goat_taxonomic_effects$effect_score[match(family, goat_taxonomic_effects$family)],
  )
```

#### - calculate susceptibility scores

1.  calculate "susceptibility effect scores" for habitat, traits, taxonomy
2.  sum traits, taxonomy susceptibilities
3.  scale total traits, taxonomy susceptibility
4.  combine with habitat susceptibility (product of two)
5.  assign susceptibility bins, confidence bins to each taxon

```{r, echo=FALSE, message=FALSE, warning=FALSE}
 goat_bin_thresholds <- c(
      "5_extremely_high" = 0.85,
      "4_high"           = 0.7,
      "3_moderate"       = 0.5,
      "2_low"            = 0.35,
      "1_no_susceptibility"        = 0
    )

# calculate "susceptibility effect scores" for habitat, traits, taxonomy
goats_assigned_susceptibility <- goats_assigned_susceptibility %>%
   # calculate confidence in susceptibility score, based on how much data available, weighted
    # by trait effect size.
      calculate_prop_traits_scored_scaled(trait_table = goat_trait_effects) %>%

  mutate(
      # confidence drops to 0.15*prop_traits_scored (LOW) if no habitat data
        susceptibility_confidence = ifelse(is.na(goat_habitat_effect_score),
                                 0.15*prop_traits_scored, 0.6*prop_traits_scored),

      # default habitat_susceptibility is set to is 0.5
        habitat_susceptibility = ifelse(is.na(goat_habitat_effect_score), .5, goat_habitat_effect_score),

      # add taxonomic effect scores
      # taxonomic effect scores are a proxy for suites of traits that are strongly linked to taxonomy
        taxonomic_susceptibility = calculate_susceptibility("taxonomic_susceptibility", taxonomic_susceptibility, .)
  ) %>%

    # add {trait}_susceptibility for every trait in goat_trait_effects (plant height, resprouting
    # capacity, bud bank location, life history, bark thickness, forb, reproductive
    # maturity, woodiness
    add_trait_susceptibility(goat_trait_effects) %>%

    mutate(
    # sum together trait and taxonomic effect scores
    # taxonomic_susceptibility weighted 3x a normal trait
      total_trait_effect_score = rowSums(across(
        c(plant_height_4m_peak_scaled_susceptibility, resprouting_capacity_susceptibility, bud_bank_location_scaled_susceptibility, 
          forb_susceptibility, life_history_susceptibility, bark_thickness_scaled_susceptibility, reproductive_maturity_scaled_susceptibility, 
          woodiness_susceptibility)), na.rm = TRUE) + 
        3*coalesce(taxonomic_susceptibility, 0),

    # rescale trait susceptibility
    # decide on range by looking at distribution of values and deciding, 
    # how much traits are "allowed" to offset habitat susceptibility
      trait_effect_score_scaled = sigmoidal_rescale(total_trait_effect_score, to = c(-0.1, 1.2)),

    # multiple scaled trait effect score susceptibility by habitat susceptibility
      total_individual_susceptibility = habitat_susceptibility*trait_effect_score_scaled,

    # NOT USING THIS. VERY POOR PREDICTIVE POWER BASED ON MODEL VALIDATION
    # For species where independent data streams identify a species as having high or low
    # goat palatability, the trait-modeled susceptibility is adjusted: 
    # - Plants that are unpalatable (plant_goat_palatability < 3), their susceptibility is *0.5
    # - Plants that are palatable (plant_goat_palatability > 3), their susceptibility is *1.5
    # - Plants that don't have independent palatability data, their susceptibility is *1
      total_individual_susceptibility_overwrite = case_when(
        is.na(plant_goat_palatability) ~ 1*total_individual_susceptibility,
        plant_goat_palatability < 3 ~ 0.75*total_individual_susceptibility,
        plant_goat_palatability >= 3 ~ 1.5*total_individual_susceptibility
        )
    )

  goats_assigned_susceptibility <- goats_assigned_susceptibility %>%

    # merge in non-threat components
    left_join(precariousness |> select(-threatened_status)) |>
    left_join(regeneration_ability |> select(-threatened_status)) |>

    # for precariousness and regeneration, assign mean value for missing data
    mutate(
      precariousness_susceptibility = ifelse(is.na(precariousness_susceptibility),
                                   mean(precariousness_susceptibility, na.rm = TRUE),
                                   precariousness_susceptibility),
      
      regeneration_ability_susceptibility = ifelse(is.na(regeneration_ability_susceptibility),
                                   mean(regeneration_ability_susceptibility, na.rm = TRUE),
                                   regeneration_ability_susceptibility)
    )

  # rescale individual susceptibility to 0-1 across all taxa before combining with
  # regeneration/precariousness, so those two terms (fixed to a narrow +/-0.1-ish
  # range everywhere) have a consistent relative influence across every threat,
  # rather than one that depends on this threat's own raw susceptibility spread
  goat_IS_range <- range(goats_assigned_susceptibility$total_individual_susceptibility, na.rm = TRUE)

  goats_assigned_susceptibility <- goats_assigned_susceptibility %>%
    mutate(
      total_individual_susceptibility_scaled = (total_individual_susceptibility - goat_IS_range[1]) / diff(goat_IS_range)
    ) %>%

    # sum together 3 components, normalise
    mutate(
      population_susceptibility = (1*total_individual_susceptibility_scaled) + regeneration_ability_susceptibility + precariousness_susceptibility,
      population_susceptibility = ifelse(total_individual_susceptibility == 0, 0, population_susceptibility)
    )

  # linearly rescale population_susceptibility to 0-1 across taxa
  goat_susceptibility_range <- range(goats_assigned_susceptibility$population_susceptibility, na.rm = TRUE)

  goats_assigned_susceptibility <- goats_assigned_susceptibility %>%
    mutate(
      population_susceptibility = (population_susceptibility - goat_susceptibility_range[1]) / diff(goat_susceptibility_range),

    # assign susceptibility bin (individual_susceptibility_bin removed - not reported to DCCEEW)
      population_susceptibility_bin = create_susceptibility_bins(population_susceptibility, bin_thresholds = goat_bin_thresholds)
    ) |>

    # confidence-based plausible range around the numeric susceptibility score
    add_susceptibility_range(
      susceptibility_col = "population_susceptibility",
      confidence_col = "susceptibility_confidence",
      bin_thresholds = goat_bin_thresholds
    )

  goats_assigned_susceptibility_threatened <- goats_assigned_susceptibility %>%
    filter(!is.na(threatened_status)) %>%
    select(taxon_name, genus, family, threatened_status,
           plant_goat_palatability,
           total_individual_susceptibility,
           population_susceptibility, population_susceptibility_bin, susceptibility_confidence,
           population_susceptibility_min, population_susceptibility_max,
           population_susceptibility_bin_min, population_susceptibility_bin_max,
           habitat, habitat_susceptibility, trait_effect_score_scaled,
           taxonomic_susceptibility, prop_traits_scored,
           contains(c("susceptibility_effect_score")), everything()) %>%
    arrange(-population_susceptibility) |>
    write_csv("~/GitHub/threatened_species_traits/export/goats_susceptibility_scores.csv", na = "")
```

#### - summary table of scorings

```{r, echo=FALSE, warning=FALSE, message=FALSE}
goat_trait_effects %>%
  arrange(effect_score, trait) %>% 
  kable(caption = "Trait effect scores for goats.") %>% 
  kable_classic()

goat_taxonomic_effects %>%
  arrange(-effect_score, family) %>% 
  kable(caption = "Taxa with explicit effect scores for goats.") %>% 
  kable_classic()

habitat_scores %>%
  select(habitat = trait_value, effect_score = goats) %>%
  group_by(effect_score) %>%
  mutate(habitat = paste0(habitat, collapse = "; ")) %>%
  ungroup() %>%
  distinct() %>% 
  select(habitat, effect_score) %>%
  arrange(effect_score) %>%
  kable(caption = "Habitat effect scores for goats.") %>% 
  kable_classic()

goats_assigned_susceptibility_threatened |>
  count(population_susceptibility_bin) |>
  mutate(pct = round(100 * n / sum(n), 0)) |>
  arrange(desc(population_susceptibility_bin)) |>
  left_join(bin_thresholds_table(goat_bin_thresholds), by = "population_susceptibility_bin") |>
  kable(caption = "taxa in each population susceptibility bin", col.names = c("Population susceptibility bin", "n", "%", "Threshold range")) %>%
  kable_classic()
```

#### - susceptibility scores density plot

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=13}
df_long_scaled <- goats_assigned_susceptibility %>%
  select(`habitat susceptibility` = habitat_susceptibility,
         `trait and taxonomy susceptibility` = trait_effect_score_scaled,
         `combined individual susceptibility` = total_individual_susceptibility,
         `mean population susceptibility` = population_susceptibility,
         `min population susceptibility` = population_susceptibility_min,
         `max population susceptibility` = population_susceptibility_max,
         threatened_status) %>%
  pivot_longer(cols = c(1:6),
               names_to = "category", values_to = "susceptibility")

make_8_panel_susceptibility_density_plot(df_long_scaled, paste0("goats (sigmoidal_rescale, k = ", sigmoidal_rescale_k, ")"))
```

#### - save 4-panel report figure to PNG

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide', include=FALSE}
png("~/GitHub/threatened_species_traits/export/figures/goats_4panel.png", width = 9, height = 6.5, units = "in", res = 150)
make_4_panel_susceptibility_density_plot(
  goats_assigned_susceptibility, "goats",
  panel_cols = c("habitat_susceptibility", "trait_effect_score_scaled", "total_individual_susceptibility", "population_susceptibility"),
  panel_labels = c("habitat susceptibility", "trait and taxonomy susceptibility", "individual susceptibility", "overall population susceptibility"),
  x_min = 0,
  x_max = 1
)
dev.off()
```

# Deer

### Assign susceptibilities to all taxa

#### - amalgamate data sources

```{r, echo=FALSE, warning=FALSE, message=FALSE}
deer_trait_effects <- tribble(
  ~trait, ~effect_score,       ~notes,
  "woodiness",-0.1,             "woody species moderately less impacted",
  "life_history",-0.2,             "plants with shorter life cycles less impacted",
  #"reproductive_maturity_scaled",-0.2, "plants with shorter life cycles less impacted", # not enough data to include
  "plant_physical_defence_structures",-0.2, "plants with physical defence less impacted",
  "resprouting_capacity", -0.2, "resprouting species are more likely to regenerate, persist",
  "plant_height_4m_peak_scaled", -0.1,  "worst height is 4 m, with decreasing impact above and below that",
  "leaf_mass_per_area_scaled",-0.2,    "species with tougher leaves (thicker, more sclerophyllous) less impacted",
  "bark_thickness_scaled",-0.1,   "higher bark thickness = lower impact, but also evidence of deer preferring to rub against rough-barked species"
)

# create dataframe with relevant traits, attributes
deer_assigned_susceptibility <- native_species %>%
  left_join(habitat, by = "taxon_name") %>%
  left_join(deer_habitat, by = "taxon_name") %>%
  left_join(plant_height |> distinct(), by = "taxon_name") %>%
  left_join(leaf_mass_per_area, by = "taxon_name") %>%
  left_join(resprouting_capacity, by = "taxon_name") %>%
  left_join(life_history, by = "taxon_name") %>%
  left_join(woodiness, by = "taxon_name") %>%
  left_join(bark_thickness, by = "taxon_name") %>%
  left_join(reproductive_maturity, by = "taxon_name") %>%
  left_join(plant_physical_defence_structures, by = "taxon_name")
```

#### - calculate susceptibility scores

1.  calculate "susceptibility effect scores" for habitat, traits, taxonomy
2.  sum traits, taxonomy susceptibilities
3.  scale total traits, taxonomy susceptibility
4.  combine with habitat susceptibility (product of two)
5.  assign susceptibility bins, confidence bins to each taxon

```{r, echo=FALSE, message=FALSE, warning=FALSE}
 deer_bin_thresholds <- c(
      "5_extremely_high" = 0.75,
      "4_high"           = 0.5,
      "3_moderate"       = 0.35,
      "2_low"            = 0.2,
      "1_no_susceptibility"        = 0
    )

# calculate "susceptibility effect scores" for habitat, traits, taxonomy
deer_assigned_susceptibility <- deer_assigned_susceptibility %>%

   # calculate confidence in susceptibility score, based on how much data available, weighted
    # by trait effect size.
      calculate_prop_traits_scored_scaled(trait_table = deer_trait_effects) %>%

  mutate(
      # confidence drops to 0.15*prop_traits_scored (LOW) if no habitat data
        susceptibility_confidence = ifelse(is.na(deer_habitat_effect_score),
                                 0.15*prop_traits_scored, 0.6*prop_traits_scored),

      # default habitat_susceptibility is set to is 0.5
        habitat_susceptibility = ifelse(is.na(deer_habitat_effect_score), .5, deer_habitat_effect_score),

      # add taxonomic effect scores
      # NONE FOR DEER
  ) %>%

    # add {trait}_susceptibility for every trait in deer_trait_effects (plant height, leaf mass
    # per area, resprouting capacity, woodiness, bark thickness, life history, plant
    # physical defence structures)
    add_trait_susceptibility(deer_trait_effects) %>%

    mutate(
      # sum trait susceptibility effect scores
        total_trait_effect_score = rowSums(across(c(plant_height_4m_peak_scaled_susceptibility, leaf_mass_per_area_scaled_susceptibility, resprouting_capacity_susceptibility, woodiness_susceptibility, plant_physical_defence_structures_susceptibility, bark_thickness_scaled_susceptibility, life_history_susceptibility
                                                    )), na.rm = TRUE),

    # rescale trait susceptibility
    # decide on range by looking at distribution of values and deciding, 
    # how much traits are "allowed" to offset habitat susceptibility
        trait_effect_score_scaled = sigmoidal_rescale(total_trait_effect_score, to = c(-0.1, 1.4)),

      # multiple scaled trait effect score susceptibility by habitat susceptibility
        total_individual_susceptibility = habitat_susceptibility*trait_effect_score_scaled
  )

  deer_assigned_susceptibility <- deer_assigned_susceptibility %>%

    # merge in non-threat components
    left_join(precariousness |> select(-threatened_status)) |>
    left_join(regeneration_ability |> select(-threatened_status)) |>

    # for precariousness and regeneration, assign mean value for missing data
    mutate(
      precariousness_susceptibility = ifelse(is.na(precariousness_susceptibility),
                                   mean(precariousness_susceptibility, na.rm = TRUE),
                                   precariousness_susceptibility),
      
      regeneration_ability_susceptibility = ifelse(is.na(regeneration_ability_susceptibility),
                                   mean(regeneration_ability_susceptibility, na.rm = TRUE),
                                   regeneration_ability_susceptibility)
    )

  # rescale individual susceptibility to 0-1 across all taxa before combining with
  # regeneration/precariousness, so those two terms (fixed to a narrow +/-0.1-ish
  # range everywhere) have a consistent relative influence across every threat,
  # rather than one that depends on this threat's own raw susceptibility spread
  deer_IS_range <- range(deer_assigned_susceptibility$total_individual_susceptibility, na.rm = TRUE)

  deer_assigned_susceptibility <- deer_assigned_susceptibility %>%
    mutate(
      total_individual_susceptibility_scaled = (total_individual_susceptibility - deer_IS_range[1]) / diff(deer_IS_range)
    ) %>%

    # sum together 3 components, normalise
    mutate(
      population_susceptibility = (1*total_individual_susceptibility_scaled) + regeneration_ability_susceptibility + precariousness_susceptibility,
      #population_susceptibility = sigmoidal_rescale(total_trait_effect_score, to = c(0.0, 1)),
      population_susceptibility = ifelse(total_individual_susceptibility == 0, 0, population_susceptibility)
    )

  # linearly rescale population_susceptibility to 0-1 across taxa
  deer_susceptibility_range <- range(deer_assigned_susceptibility$population_susceptibility, na.rm = TRUE)

  deer_assigned_susceptibility <- deer_assigned_susceptibility %>%
    mutate(
      population_susceptibility = (population_susceptibility - deer_susceptibility_range[1]) / diff(deer_susceptibility_range),

    # assign susceptibility bin (individual_susceptibility_bin removed - not reported to DCCEEW)
      population_susceptibility_bin = create_susceptibility_bins(population_susceptibility,
                                             bin_thresholds = deer_bin_thresholds)
    ) |>

    # confidence-based plausible range around the numeric susceptibility score
    add_susceptibility_range(
      susceptibility_col = "population_susceptibility",
      confidence_col = "susceptibility_confidence",
      bin_thresholds = deer_bin_thresholds
    )

  deer_assigned_susceptibility_threatened <- deer_assigned_susceptibility %>%
    filter(!is.na(threatened_status)) %>%
    mutate(taxonomic_susceptibility = 0) %>%
    select(taxon_name, genus, family, threatened_status,
           total_individual_susceptibility,
           population_susceptibility, population_susceptibility_bin, susceptibility_confidence,
           population_susceptibility_min, population_susceptibility_max,
           population_susceptibility_bin_min, population_susceptibility_bin_max,
           habitat, habitat_susceptibility, trait_effect_score_scaled,
           taxonomic_susceptibility, prop_traits_scored,
           contains(c("susceptibility_effect_score")), everything()) %>%
    arrange(-population_susceptibility) |>
    write_csv("~/GitHub/threatened_species_traits/export/deer_susceptibility_scores.csv", na = "")
```

#### - summary table of scorings

```{r, echo=FALSE, warning=FALSE, message=FALSE}
deer_trait_effects %>%
  arrange(trait, effect_score) %>% 
  kable(caption = "Trait effect scores for deer") %>% 
  kable_classic()

habitat_scores %>%
  select(habitat = trait_value, effect_score = deer) %>%
  group_by(effect_score) %>%
  mutate(habitat = paste0(habitat, collapse = "; ")) %>%
  ungroup() %>%
  distinct() %>% 
  select(habitat, effect_score) %>%
  arrange(effect_score) %>%
  kable(caption = "Habitat scores for deer.") %>% 
  kable_classic()

deer_assigned_susceptibility_threatened |>
  count(population_susceptibility_bin) |>
  mutate(pct = round(100 * n / sum(n), 0)) |>
  arrange(desc(population_susceptibility_bin)) |>
  left_join(bin_thresholds_table(deer_bin_thresholds), by = "population_susceptibility_bin") |>
  kable(caption = "taxa in each population susceptibility bin", col.names = c("Population susceptibility bin", "n", "%", "Threshold range")) %>%
  kable_classic()
```

#### - susceptibility scores density plot

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=13}
df_long_scaled <- deer_assigned_susceptibility %>%
  select(`habitat susceptibility` = habitat_susceptibility,
         `trait and taxonomy susceptibility` = trait_effect_score_scaled,
         `combined individual susceptibility` = total_individual_susceptibility,
         `mean population susceptibility` = population_susceptibility,
         `min population susceptibility` = population_susceptibility_min,
         `max population susceptibility` = population_susceptibility_max,
         threatened_status) %>%
  pivot_longer(cols = c(1:6),
               names_to = "category", values_to = "susceptibility")

make_8_panel_susceptibility_density_plot(df_long_scaled, paste0("deer (sigmoidal_rescale, k = ", sigmoidal_rescale_k, ")"))
```

#### - save 4-panel report figure to PNG

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide', include=FALSE}
png("~/GitHub/threatened_species_traits/export/figures/deer_4panel.png", width = 9, height = 6.5, units = "in", res = 150)
make_4_panel_susceptibility_density_plot(
  deer_assigned_susceptibility, "deer",
  panel_cols = c("habitat_susceptibility", "trait_effect_score_scaled", "total_individual_susceptibility", "population_susceptibility"),
  panel_labels = c("habitat susceptibility", "trait susceptibility", "individual susceptibility", "overall population susceptibility"),
  x_min = 0,
  x_max = 1.35
)
dev.off()
```

# Pigs

### Assign susceptibilities to all taxa

#### - amalgamate data sources

```{r, echo=FALSE, message=FALSE, warning=FALSE}
pigs_trait_effects <- tribble(
  ~trait, ~effect_score,       ~notes,
  "woodiness",-0.4,             "woody species less impacted",
  "resprouting_capacity", -0.2, "resprouting species are more likely to regenerate, persist",
  "plant_height_scaled", -0.2,  "shorter plants more impacted by pigs",
  "fruit_fleshiness",NA,       "fleshy fruit = higher impact, because preferred food source; merged with fruit_length for scoring",
  "fruit_length_scaled",0.25,    "large fruit = higher impact, because preferred food source; incorporates fruit_fleshiness for scoring, because length doesn't matter if not fleshy",
  "storage_organ_fleshy",0.4,   "fleshy underground storage organ = much higher impact, because single most preferred food source",
  "plant_growth_substrate_pigs",0.4, "epiphytes and lithophytes at lower susceptibility than terrestrial or aquatic species"
)

# For families & genera known to have fleshy underground storage organs or
# other pig-preferred traits, all species are assigned a taxonomic_effect_score = 0.4
pigs_taxonomic_vector <- c("Goodeniaceae", "Orchidaceae", "Cyperaceae", "Dioscoreaceae", "Iridaceae", "Hydrocharitaceae", "Marsileaceae", "Nymphaeaceae", "Pontederiaceae", "Restionaceae", "Eriocaulaceae", "Thismiaceae", "Amaryllidaceae", "Alismataceae", "Apiaceae", "Apocynaceae", "Aponogetonaceae", "Araliaceae", "Arecaceae", "Aspleniaceae", "Athyriaceae", "Convolvulaceae", "Cyatheaceae", "Musaceae", "Zingiberaceae", "Utricularia", "Ipomoea", "Ampelocissus", "Bulbine", "Microseris", "Sonchus")

pigs_taxonomic_effects <- as.data.frame(pigs_taxonomic_vector) %>%
  rename(taxon = pigs_taxonomic_vector) %>%
  mutate( 
    taxon_rank = ifelse(str_detect(taxon, "aceae"), "family", "genus"),
    effect_score = 0.4
    )

# create dataframe with relevant traits, attributes
pigs_assigned_susceptibility <- native_species %>%
  left_join(habitat, by = "taxon_name") %>%
  left_join(pigs_habitat, by = "taxon_name") %>%
  left_join(plant_height |> distinct(), by = "taxon_name") %>%
  left_join(resprouting_capacity, by = "taxon_name") %>%
  left_join(life_history, by = "taxon_name") %>%
  left_join(storage_organ_fleshy, by = "taxon_name") %>%
  left_join(fruit_fleshiness, by = "taxon_name") %>%
  left_join(fruit_length, by = "taxon_name") %>%
  left_join(woodiness, by = "taxon_name") %>%
  left_join(plant_growth_substrate_pigs, by = "taxon_name") |>
  # needed because most terrestrial plants unscored; trait well-filled for aquatic, epiphyte, lithophyte, but not terrestrial
  mutate(plant_growth_substrate_pigs = ifelse(is.na(plant_growth_substrate_pigs), 0, plant_growth_substrate_pigs)) |>
  left_join(pigs_taxonomic_effects %>% filter(taxon_rank == "family") %>% rename(family = taxon) %>% select(-taxon_rank) %>% rename(taxonomic_effect = effect_score), by = "family") %>%
  left_join(pigs_taxonomic_effects %>% filter(taxon_rank == "genus") %>% rename(genus = taxon) %>% select(-taxon_rank) %>% rename(taxonomic_effect2 = effect_score), by = "genus") %>%
  mutate(taxonomic_effect = ifelse(!is.na(taxonomic_effect2), taxonomic_effect2, taxonomic_effect)) %>%
  select(-taxonomic_effect2) %>%
  # fruit_length_scaled is multiplied by fruit_fleshiness (when known), because fruit
  # dimensions only matter to pigs if the fruit is fleshy
  mutate(fruit_length_scaled = ifelse(!is.na(fruit_fleshiness), fruit_length_scaled*fruit_fleshiness, fruit_length_scaled)) # b/c fruit dimensions only matter for fleshy fruits
```

#### - calculate susceptibility scores

1.  calculate "susceptibility effect scores" for habitat, traits, taxonomy
2.  sum traits, taxonomy susceptibilities
3.  scale total traits, taxonomy susceptibility
4.  combine with habitat susceptibility (product of two)
5.  assign susceptibility bins, confidence bins to each taxon

```{r, echo=FALSE, message=FALSE, warning=FALSE}
 pig_bin_thresholds <- c(
      "5_extremely_high" = 0.95,
      "4_high"           = 0.7,
      "3_moderate"       = 0.5,
      "2_low"            = 0.3,
      "1_no_susceptibility"        = 0
    )
 
# calculate "susceptibility effect scores" for habitat, traits, taxonomy
pigs_assigned_susceptibility <- pigs_assigned_susceptibility %>%
   # calculate confidence in susceptibility score, based on how much data available, weighted
    # by trait effect size.
      calculate_prop_traits_scored_scaled(trait_table = pigs_trait_effects) %>%

  mutate(
      # confidence drops to 0.15*prop_traits_scored if no habitat data
        susceptibility_confidence = ifelse(is.na(pig_habitat_effect_score),
                                 0.15*prop_traits_scored, 0.6*prop_traits_scored),

      # default habitat_susceptibility is set to is 0.5
        habitat_susceptibility = ifelse(is.na(pig_habitat_effect_score), .5, pig_habitat_effect_score),

      # add taxonomic effect scores
      # taxonomic effect scores are a proxy for suites of traits that are strongly linked to taxonomy
        taxonomic_susceptibility = calculate_susceptibility("taxonomic_effect", taxonomic_effect, .)
  ) %>%

    # add {trait}_susceptibility for every trait in pigs_trait_effects (plant height, woodiness,
    # resprouting capacity, storage organ, fruit length
    add_trait_susceptibility(pigs_trait_effects) %>%

    mutate(
      # sum together trait and taxonomic effect scores
        # taxonomic_susceptibility weighted 3x a normal trait
        total_trait_effect_score = rowSums(across(
          c(plant_height_scaled_susceptibility, woodiness_susceptibility, fruit_length_scaled_susceptibility, 
            resprouting_capacity_susceptibility, storage_organ_fleshy_susceptibility, plant_growth_substrate_pigs_susceptibility)), na.rm = TRUE) + 
          3*coalesce(taxonomic_susceptibility, 0),

    # rescale trait susceptibility
    # decide on range by looking at distribution of values and deciding, 
    # how much traits are "allowed" to offset habitat susceptibility
        trait_effect_score_scaled = sigmoidal_rescale(total_trait_effect_score, to = c(-0.1, 1.3)),

      # multiple scaled trait effect score susceptibility by habitat susceptibility
        total_individual_susceptibility = habitat_susceptibility*trait_effect_score_scaled
  )

  pigs_assigned_susceptibility <- pigs_assigned_susceptibility %>%

    # merge in non-threat components
    left_join(precariousness |> select(-threatened_status)) |>
    left_join(regeneration_ability |> select(-threatened_status)) |>

    # for precariousness and regeneration, assign mean value for missing data
    mutate(
      precariousness_susceptibility = ifelse(is.na(precariousness_susceptibility),
                                   mean(precariousness_susceptibility, na.rm = TRUE),
                                   precariousness_susceptibility),
      
      regeneration_ability_susceptibility = ifelse(is.na(regeneration_ability_susceptibility),
                                   mean(regeneration_ability_susceptibility, na.rm = TRUE),
                                   regeneration_ability_susceptibility)
    )

  # rescale individual susceptibility to 0-1 across all taxa before combining with
  # regeneration/precariousness, so those two terms (fixed to a narrow +/-0.1-ish
  # range everywhere) have a consistent relative influence across every threat,
  # rather than one that depends on this threat's own raw susceptibility spread
  pig_IS_range <- range(pigs_assigned_susceptibility$total_individual_susceptibility, na.rm = TRUE)

  pigs_assigned_susceptibility <- pigs_assigned_susceptibility %>%
    mutate(
      total_individual_susceptibility_scaled = (total_individual_susceptibility - pig_IS_range[1]) / diff(pig_IS_range)
    ) %>%

    # sum together 3 components, normalise
    mutate(
      population_susceptibility = (1*total_individual_susceptibility_scaled) + regeneration_ability_susceptibility + precariousness_susceptibility,
      population_susceptibility = ifelse(total_individual_susceptibility == 0, 0, population_susceptibility)
    )

  # linearly rescale population_susceptibility to 0-1 across taxa
  pig_susceptibility_range <- range(pigs_assigned_susceptibility$population_susceptibility, na.rm = TRUE)

  pigs_assigned_susceptibility <- pigs_assigned_susceptibility %>%
    mutate(
      population_susceptibility = (population_susceptibility - pig_susceptibility_range[1]) / diff(pig_susceptibility_range),

    # assign susceptibility bin (individual_susceptibility_bin removed - not reported to DCCEEW)
      population_susceptibility_bin = create_susceptibility_bins(population_susceptibility, bin_thresholds = pig_bin_thresholds)
    ) |>

    # confidence-based plausible range around the numeric susceptibility score
    add_susceptibility_range(
      susceptibility_col = "population_susceptibility",
      confidence_col = "susceptibility_confidence",
      bin_thresholds = pig_bin_thresholds
    )


  pigs_assigned_susceptibility_threatened <- pigs_assigned_susceptibility |>
    filter(!is.na(threatened_status)) |>
    mutate(taxonomic_susceptibility = 0) |>
    select(taxon_name, genus, family, threatened_status,
           total_individual_susceptibility,
           total_individual_susceptibility,
           population_susceptibility, population_susceptibility_bin, susceptibility_confidence,
           population_susceptibility_min, population_susceptibility_max,
           population_susceptibility_bin_min, population_susceptibility_bin_max,
           habitat, habitat_susceptibility, trait_effect_score_scaled,
           taxonomic_susceptibility, prop_traits_scored,
           contains(c("susceptibility_effect_score")), everything()) %>%
    arrange(-population_susceptibility) |>
    write_csv("~/GitHub/threatened_species_traits/export/pigs_susceptibility_scores.csv", na = "")
```

#### - summary table of scorings

```{r, echo=FALSE, warning=FALSE, message=FALSE}
pigs_trait_effects %>%
  arrange(effect_score, trait) %>% 
  kable(caption = "Trait effect scores for pigs.") %>% 
  kable_classic()

pigs_taxonomic_effects %>%
  arrange(-effect_score, taxon_rank, taxon) %>% 
  kable(caption = "Taxa with explicit taxonomic effect scores for pigs.") %>% 
  kable_classic()

habitat_scores %>%
  select(habitat = trait_value, effect_score = pigs) %>%
  group_by(effect_score) %>%
  mutate(habitat = paste0(habitat, collapse = "; ")) %>%
  ungroup() %>%
  distinct() %>% 
  select(habitat, effect_score) %>%
  arrange(effect_score) %>%
  kable(caption = "Habitat effect scores for pigs.") %>% 
  kable_classic()

pigs_assigned_susceptibility_threatened |>
  count(population_susceptibility_bin) |>
  mutate(pct = round(100 * n / sum(n), 0)) |>
  arrange(desc(population_susceptibility_bin)) |>
  left_join(bin_thresholds_table(pig_bin_thresholds), by = "population_susceptibility_bin") |>
  kable(caption = "taxa in each population susceptibility bin", col.names = c("Population susceptibility bin", "n", "%", "Threshold range")) %>%
  kable_classic()
```

#### - susceptibility scores density plot

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=13}
df_long_scaled <- pigs_assigned_susceptibility %>%
  select(`habitat susceptibility` = habitat_susceptibility,
         `trait and taxonomy susceptibility` = trait_effect_score_scaled,
         `combined individual susceptibility` = total_individual_susceptibility,
         `mean population susceptibility` = population_susceptibility,
         `min population susceptibility` = population_susceptibility_min,
         `max population susceptibility` = population_susceptibility_max,
         threatened_status) %>%
  pivot_longer(cols = c(1:6),
               names_to = "category", values_to = "susceptibility")

make_8_panel_susceptibility_density_plot(df_long_scaled, paste0("pigs (sigmoidal_rescale, k = ", sigmoidal_rescale_k, ")"))
```

#### - save 4-panel report figure to PNG

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide', include=FALSE}
png("~/GitHub/threatened_species_traits/export/figures/pigs_4panel.png", width = 9, height = 6.5, units = "in", res = 150)
make_4_panel_susceptibility_density_plot(
  pigs_assigned_susceptibility, "pigs",
  panel_cols = c("habitat_susceptibility", "trait_effect_score_scaled", "total_individual_susceptibility", "population_susceptibility"),
  panel_labels = c("habitat susceptibility", "trait and taxonomy susceptibility", "individual susceptibility", "overall population susceptibility"),
  x_min = 0,
  x_max = 1.2
)
dev.off()
```

# C4 tropical grasses

### Assign susceptibilities to all taxa

#### - amalgamate data sources

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tropical_C4_grasses_trait_effects <- tribble(
  ~trait,                  ~effect_score, ~notes,
  "woodiness",              0.3,          "woody taxa = higher impact",
  "life_history",           0.2,          "annuals, ephemerals = lower impact",
  "plant_height_scaled",   -0.1,          "taller plants = lower impact, because plants are taller than invasives",
  "resprouting_capacity",  -0.2,          "resprouters = slower impact, with respect to increased fire frequency",
  "storage_organ_present", -0.3,          "includes geophytes",
  "bud_bank_location_scaled",     -0.1,         "used twice, high bud banks a negative for increased fire intensity, a positive for high fire frequency; offsets half the benefit of being a resprouter",
  "reproductive_maturity_scaled", -0.3,          "plants with a shorter juvenile period more likely to set seed before next fire",
  "dry_season_flowering", NA, "scored directly with two separate coefficients: -0.3 in direct_competition_susceptibility (competition benefit), -0.1 in fire_season_susceptibility (fire cost)"

)

# taxonomic vector - note this is just for the kable printout; actually scoring is done below
tropical_C4_grasses_taxonomic_vector <- c("perennial Poaceae")

tropical_C4_grasses_taxonomic_effects <- as.data.frame(tropical_C4_grasses_taxonomic_vector) %>%
  rename(taxon = tropical_C4_grasses_taxonomic_vector) %>%
  mutate(
    effect_score = 0.4
    )

# create dataframe with relevant traits, attributes
tropical_C4_grasses_assigned_susceptibility <- native_species %>%
  left_join(habitat, by = "taxon_name") %>%
  left_join(C4_grasses_habitat, by = "taxon_name") %>%
  left_join(tropical_grass_flowering_times, by = "taxon_name") %>%
  left_join(plant_height |> distinct(), by = "taxon_name") %>%
  left_join(reproductive_maturity, by = "taxon_name") %>%
  left_join(life_history, by = "taxon_name") %>%
  left_join(geophyte, by = "taxon_name") %>%
  left_join(woodiness, by = "taxon_name") %>%
  left_join(resprouting_capacity, by = "taxon_name") %>%
  left_join(bud_bank_location_scaled, by = "taxon_name") %>%
  left_join(storage_organ_present, by = "taxon_name") %>%
  # geophyte status is folded into `storage_organ_present`, retaining whichever value is higher, i.e. more "at susceptibility"
  mutate(storage_organ_present = ifelse(is.na(storage_organ_present)|storage_organ_present < geophyte, geophyte, storage_organ_present))
```

#### - calculate susceptibility scores

1.  calculate "susceptibility effect scores" for habitat, traits, taxonomy
2.  sum traits, taxonomy susceptibilities
3.  scale total traits, taxonomy susceptibility
4.  combine with habitat susceptibility (product of two)
5.  assign susceptibility bins, confidence bins to each taxon

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# assign susceptibility by category
  # C4 grasses uses different bin thresholds to every other threat - defined once here
  # and reused below for both the point estimate and the range
  tropical_C4_grasses_bin_thresholds <- c(
      "5_extremely_high" = 0.4,
      "4_high"           = 0.3,
      "3_moderate"       = 0.2,
      "2_low"            = 0.15,
      "1_no_susceptibility"        = 0
    )

# calculate "susceptibility effect scores" for habitat, traits, taxonomy
tropical_C4_grasses_assigned_susceptibility <- tropical_C4_grasses_assigned_susceptibility %>%
   # calculate confidence in susceptibility score, based on how much data available, weighted
    # by trait effect size.
      calculate_prop_traits_scored_scaled(trait_table = tropical_C4_grasses_trait_effects) %>%

  mutate(
      # confidence drops to 0.15*prop_traits_scored if no habitat data
        susceptibility_confidence = ifelse(is.na(C4_grasses_habitat_effect_score),
                                 0.15*prop_traits_scored, 0.6*prop_traits_scored),

      # when available use habitat susceptibility based on species habitat; default habitat = 0.5
        habitat_susceptibility = ifelse(is.na(C4_grasses_habitat_effect_score),0.5,
                              C4_grasses_habitat_effect_score)
  ) %>%

    # add {trait}_susceptibility for every trait in tropical_C4_grasses_trait_effects (plant
    # height, storage organ, reproductive maturity, life history, woodiness,
    # resprouting capacity, bud bank location). dry_season_flowering is deliberately
    # NOT in this tribble - see 27/28 below, where it's scored with two different,
    # expert-elicited coefficients instead of a single trait-effects entry
    add_trait_susceptibility(tropical_C4_grasses_trait_effects) %>%

    mutate(
      # sign-flipped reuse of the same trait susceptibility: a high bud bank is protective against fire intensity,
      # but is treated as a slight liability for fire frequency
      bud_bank_location_scaled_susceptibility2 = -bud_bank_location_scaled_susceptibility,

    # taxonomy effect score

    # for this threat, the taxonomy_effect_score only applies to perennial grasses:
    # life_history > 1.8 & family == "Poaceae"
      taxonomy_effect_score = ifelse(!is.na(life_history) & life_history > 1.8 & family == "Poaceae", 0.4, 0),

    # combine trait susceptibility effect scores

    # dry_season_flowering: less direct competition if 1) you are tall or 2) you are an annual,
    # especially ones that grows outside the C4 grass growing season
      direct_competition_susceptibility = ifelse(!is.na(plant_height) | (!is.na(dry_season_flowering) & life_history < 1.5),
                                       -0.3*dry_season_flowering + plant_height_scaled_susceptibility + life_history_susceptibility,
                                       NA),

    # rescue step: the formula above can still come out NA even when its own condition
    # matched - e.g. plant_height known but dry_season_flowering unknown, so
    # -0.3*dry_season_flowering poisons the whole sum with NA via arithmetic propagation.
    # When that happens and plant_height is known, fall back to plant_height_scaled_susceptibility
    # alone rather than losing the signal entirely to the later blanket NA->0 replacement.
    # Species with neither plant_height nor a qualifying dry_season_flowering get 0.3 (default).
      direct_competition_susceptibility = ifelse(!is.na(plant_height) & is.na(direct_competition_susceptibility),
                                       plant_height_scaled_susceptibility,
                                       ifelse(is.na(direct_competition_susceptibility), 0.3, direct_competition_susceptibility)),

    # most impacted by increased fire intensity if you are perennial, woody and with high bud bank location; having a storage organ offers a bit of reprieve
      fire_intensity_susceptibility = rowSums(across(c(woodiness_susceptibility, storage_organ_present_susceptibility, bud_bank_location_scaled_susceptibility, life_history_susceptibility)), na.rm = TRUE),
    
    # most impacted by increased fire frequency if you are an obligate seeder or with basal buds; and being perennial, especially with long time to maturity
    # reproductive_maturity_susceptibility, should be in here, but so little data it will skew results
      fire_frequency_susceptibility = rowSums(across(c(life_history_susceptibility, resprouting_capacity_susceptibility, bud_bank_location_scaled_susceptibility2)), na.rm = TRUE),
    
    # fire_season_susceptibility: less impacted by fire season if you are an annual that flowers and fruits in the late dry season
      fire_season_susceptibility = ifelse(!is.na(life_history) & !is.na(dry_season_flowering) & life_history < 1.5, -0.1*dry_season_flowering, .1),
    
    # replace na's with 0
      across(c(direct_competition_susceptibility, habitat_susceptibility, fire_season_susceptibility, fire_intensity_susceptibility,fire_frequency_susceptibility, taxonomy_effect_score), ~tidyr::replace_na(.x, 0)),
    
    # determine total susceptibility, fire season "worth less", because smaller effect, direct competition largest
      # taxonomy_effect_score weighted 2x a normal trait
      total_trait_effect_score = rowSums(across(
        c(direct_competition_susceptibility, fire_season_susceptibility, fire_intensity_susceptibility, 
          fire_frequency_susceptibility)), na.rm = TRUE) + 
      3*coalesce(taxonomy_effect_score, 0),
    
    
    # rescale trait susceptibility
    # decide on range by looking at distribution of values and deciding, 
    # how much traits are "allowed" to offset habitat susceptibility
      trait_effect_score_scaled = sigmoidal_rescale(total_trait_effect_score, to = c(0.1, 1.3)),

    # multiple scaled trait effect score susceptibility by habitat susceptibility
      total_individual_susceptibility = habitat_susceptibility*trait_effect_score_scaled
  )


  tropical_C4_grasses_assigned_susceptibility <- tropical_C4_grasses_assigned_susceptibility %>%

    # merge in non-threat components
    left_join(precariousness |> select(-threatened_status)) |>
    left_join(regeneration_ability |> select(-threatened_status)) |>

    # for precariousness and regeneration, assign mean value for missing data
    mutate(
      precariousness_susceptibility = ifelse(is.na(precariousness_susceptibility),
                                   mean(precariousness_susceptibility, na.rm = TRUE),
                                   precariousness_susceptibility),
      
      regeneration_ability_susceptibility = ifelse(is.na(regeneration_ability_susceptibility),
                                   mean(regeneration_ability_susceptibility, na.rm = TRUE),
                                   regeneration_ability_susceptibility)
    )

  # rescale individual susceptibility to 0-1 across all taxa before combining with
  # regeneration/precariousness, so those two terms (fixed to a narrow +/-0.1-ish
  # range everywhere) have a consistent relative influence across every threat,
  # rather than one that depends on this threat's own raw susceptibility spread
  tropical_C4_grasses_IS_range <- range(tropical_C4_grasses_assigned_susceptibility$total_individual_susceptibility, na.rm = TRUE)

  tropical_C4_grasses_assigned_susceptibility <- tropical_C4_grasses_assigned_susceptibility %>%
    mutate(
      total_individual_susceptibility_scaled = (total_individual_susceptibility - tropical_C4_grasses_IS_range[1]) / diff(tropical_C4_grasses_IS_range)
    ) %>%

    # sum together 3 components, normalise
    mutate(
      population_susceptibility = (1*total_individual_susceptibility_scaled) + regeneration_ability_susceptibility + precariousness_susceptibility,
      population_susceptibility = ifelse(total_individual_susceptibility == 0, 0, population_susceptibility)
    )

  # linearly rescale population_susceptibility to 0-1 across taxa
  tropical_C4_grasses_susceptibility_range <- range(tropical_C4_grasses_assigned_susceptibility$population_susceptibility, na.rm = TRUE)

  tropical_C4_grasses_assigned_susceptibility <- tropical_C4_grasses_assigned_susceptibility %>%
    mutate(
      population_susceptibility = (population_susceptibility - tropical_C4_grasses_susceptibility_range[1]) / diff(tropical_C4_grasses_susceptibility_range),

    # assign susceptibility bin (individual_susceptibility_bin removed - not reported to DCCEEW)
      population_susceptibility_bin = create_susceptibility_bins(population_susceptibility,
        bin_thresholds = tropical_C4_grasses_bin_thresholds)
    ) |>

    # confidence-based plausible range around the numeric susceptibility score
    add_susceptibility_range(
      susceptibility_col = "population_susceptibility",
      confidence_col = "susceptibility_confidence",
      bin_thresholds = tropical_C4_grasses_bin_thresholds
    )

  tropical_C4_grasses_assigned_susceptibility_threatened <- tropical_C4_grasses_assigned_susceptibility |>
    filter(!is.na(threatened_status)) |>
    mutate(taxonomic_susceptibility = 0) |>
    # coding exercise (2026-08-24): flag-only column, not used in scoring - see
    # c4_grasses_known_impact_species above
    mutate(known_or_suspected_impact = ifelse(taxon_name %in% c4_grasses_known_impact_species, "X", NA)) |>
    select(taxon_name, genus, family, threatened_status,
           known_or_suspected_impact,
           total_individual_susceptibility,
           population_susceptibility, population_susceptibility_bin, susceptibility_confidence,
           population_susceptibility_min, population_susceptibility_max,
           population_susceptibility_bin_min, population_susceptibility_bin_max,
           habitat, habitat_susceptibility, trait_effect_score_scaled,
           taxonomic_susceptibility, prop_traits_scored,
           direct_competition_susceptibility, fire_intensity_susceptibility,
           fire_frequency_susceptibility, fire_season_susceptibility,
           contains(c("susceptibility_effect_score")), everything()) |>
    arrange(-population_susceptibility) |>
    write_csv("~/GitHub/threatened_species_traits/export/tropical_C4_grasses_susceptibility_scores.csv", na = "")
```

#### - summary table of scorings

```{r, echo=FALSE, warning=FALSE, message=FALSE}
tropical_C4_grasses_trait_effects %>%
  arrange(effect_score, trait) %>% 
  kable(caption = "Trait effect scores for invasive tropical C4 grasses.") %>% 
  kable_classic()

tropical_C4_grasses_taxonomic_effects %>%
  arrange(-effect_score, taxon) %>% 
  kable(caption = "Taxa with explicit taxonomic effect scores for invasive tropical C4 grasses.") %>% 
  kable_classic()

habitat_scores %>%
  select(habitat = trait_value, effect_score = C4_grasses) %>%
  group_by(effect_score) %>%
  mutate(habitat = paste0(habitat, collapse = "; ")) %>%
  ungroup() %>%
  distinct() %>% 
  select(habitat, effect_score) %>%
  arrange(effect_score) %>%
  kable(caption = "habitat scores for C4 perennial grasses") %>% 
  kable_classic()

tropical_C4_grasses_assigned_susceptibility_threatened |>
  count(population_susceptibility_bin) |>
  mutate(pct = round(100 * n / sum(n), 0)) |>
  arrange(desc(population_susceptibility_bin)) |>
  left_join(bin_thresholds_table(tropical_C4_grasses_bin_thresholds), by = "population_susceptibility_bin") |>
  kable(caption = "taxa in each population susceptibility bin", col.names = c("Population susceptibility bin", "n", "%", "Threshold range")) %>%
  kable_classic()
```

#### - susceptibility scores density plot

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=13}
df_long_scaled <- tropical_C4_grasses_assigned_susceptibility %>%
  select(`habitat susceptibility` = habitat_susceptibility,
         `trait and taxonomy susceptibility` = trait_effect_score_scaled,
         `combined individual susceptibility` = total_individual_susceptibility,
         `mean population susceptibility` = population_susceptibility,
         `min population susceptibility` = population_susceptibility_min,
         `max population susceptibility` = population_susceptibility_max,
         threatened_status) %>%
  pivot_longer(cols = c(1:6),
               names_to = "category", values_to = "susceptibility")

make_8_panel_susceptibility_density_plot(df_long_scaled, paste0("C4 grasses (sigmoidal_rescale, k = ", sigmoidal_rescale_k, ")"))
```

#### - save 4-panel report figure to PNG

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide', include=FALSE}
png("~/GitHub/threatened_species_traits/export/figures/C4_grasses_4panel.png", width = 9, height = 6.5, units = "in", res = 150)
make_4_panel_susceptibility_density_plot(
  tropical_C4_grasses_assigned_susceptibility, "C4 grasses",
  panel_cols = c("habitat_susceptibility", "trait_effect_score_scaled", "total_individual_susceptibility", "population_susceptibility"),
  panel_labels = c("habitat susceptibility", "trait and taxonomy susceptibility", "individual susceptibility", "overall population susceptibility"),
  x_min = 0, x_max = 1.5
)
dev.off()
```

# Hymenachne

### Assign susceptibilities to all taxa

#### - amalgamate data sources

```{r, echo=FALSE, warning=FALSE, message=FALSE}
hymenachne_trait_effects <- tribble(
  ~trait, ~effect_score, ~notes,
  "woodiness", -0.4, "woody taxa = lower impact",
  "life_history", -0.3, "annuals, ephemerals = higher impact",
  "plant_height_scaled", -0.3, "taller plants = lower impact",
  "pollination_syndrome", -0.3, "wind pollinated species more persistent",
  "plant_growth_substrate_hymenanche", 2, "aquatic/semiaquatic species = much higher impact"
)

# 14 families that are associated with aquatic and wetland environments are identified
# as having a higher susceptibility from hymenachne
# All species in these families are assigned a taxonomic effect score
hymenachne_taxonomic_vector_old <- 
  c("Arecaceae", "Cyperaceae", "Haloragaceae", "Hydrocharitaceae", "Iridaceae", 
  "Juncaceae", "Marsileaceae", "Nymphaeaceae", "Orchidaceae", "Poaceae", 
  "Pontederiaceae", "Restionaceae", "Typhaceae")

hymenachne_taxonomic_vector <- 
  c("Arecaceae", "Haloragaceae", "Hydrocharitaceae",
  "Juncaceae", "Marsileaceae", "Nymphaeaceae", 
  "Pontederiaceae", "Restionaceae", "Typhaceae")

hymenachne_taxonomic_effects <- as.data.frame(hymenachne_taxonomic_vector) %>%
  rename(taxon = hymenachne_taxonomic_vector) %>%
  mutate(
    effect_score = 0.4
    )

# create dataframe with relevant traits, attributes
hymenachne_assigned_susceptibility <- native_species %>%
  left_join(habitat, by = "taxon_name") %>%
  left_join(plant_growth_substrate_hymenanche, by = "taxon_name") %>%
  left_join(hymenachne_habitat, by = "taxon_name") %>%
  left_join(plant_height |> distinct(), by = "taxon_name") %>%
  left_join(life_history, by = "taxon_name") %>%
  left_join(woodiness, by = "taxon_name") %>%
  left_join(pollination_syndrome, by = "taxon_name") %>%
  mutate(
    taxonomic_susceptibility = ifelse(family %in% hymenachne_taxonomic_effects$taxon,
                            0.4, 0))
```

#### - calculate susceptibility scores

1.  calculate "susceptibility effect scores" for habitat, traits, taxonomy
2.  sum traits, taxonomy susceptibilities
3.  scale total traits, taxonomy susceptibility
4.  combine with habitat susceptibility (product of two)
5.  assign susceptibility bins, confidence bins to each taxon

```{r, echo=FALSE, message=FALSE, warning=FALSE}
 hymenachne_bin_thresholds <- c(
      "5_extremely_high" = 0.8,
      "4_high"           = 0.6,
      "3_moderate"       = 0.4,
      "2_low"            = 0.1,
      "1_no_susceptibility"        = 0
    )

# calculate "susceptibility effect scores" for habitat, traits, taxonomy
hymenachne_assigned_susceptibility <- hymenachne_assigned_susceptibility %>%
  
   # calculate confidence in susceptibility score, based on how much data available, weighted
    # by trait effect size.
      calculate_prop_traits_scored_scaled(trait_table = hymenachne_trait_effects) %>%

  mutate(
    # habitat_susceptibility is just the habitat score. When habitat is missing (a lot of
    # missing habitat scores were producing weird/uninformative responses),
    # fall back to the plant_growth_substrate_hymenanche value (0/1, aquatic vs
    # not) as a proxy for habitat susceptibility; only default to 0.3 when neither is available.
    # plant_growth_substrate is separately also scored as a trait (via
    # add_trait_susceptibility below, effect_score = 1).
      habitat_susceptibility = case_when(
        !is.na(hymenachne_habitat_effect_score) & hymenachne_habitat_effect_score != 0 ~ hymenachne_habitat_effect_score,
        !is.na(plant_growth_substrate_hymenanche) ~ plant_growth_substrate_hymenanche,
        !is.na(hymenachne_habitat_effect_score) ~ hymenachne_habitat_effect_score,
        TRUE ~ 0.3
      ),

    # confidence drops to 0.15*prop_traits_scored if there's no habitat score
    # and no growth substrate fallback either
        susceptibility_confidence = ifelse(
          is.na(hymenachne_habitat_effect_score) & is.na(plant_growth_substrate_hymenanche),
          0.15*prop_traits_scored, .6*prop_traits_scored),

    # add taxonomic effect scores
      taxonomic_susceptibility = calculate_susceptibility("taxonomic_susceptibility", taxonomic_susceptibility, .)
  ) %>%

    # add {trait}_susceptibility for every trait in hymenachne_trait_effects (plant height,
    # woodiness, life history, pollination syndrome, plant growth substrate)
    add_trait_susceptibility(hymenachne_trait_effects) %>%

    mutate(
    # sum trait susceptibility effect scores
      # taxonomic_susceptibility weighted 3x a normal trait
      total_trait_effect_score = rowSums(across(
        c(plant_height_scaled_susceptibility, life_history_susceptibility, woodiness_susceptibility,
          pollination_syndrome_susceptibility, plant_growth_substrate_hymenanche_susceptibility)), na.rm = TRUE) +
        1*coalesce(taxonomic_susceptibility, 0),

    # rescale trait susceptibility
    # decide on range by looking at distribution of values and deciding, 
    # how much traits are "allowed" to offset habitat susceptibility
      trait_effect_score_scaled = sigmoidal_rescale(total_trait_effect_score, to = c(0.0, 1.2)),

    # multiple scaled trait effect score susceptibility by habitat susceptibility
      total_individual_susceptibility = habitat_susceptibility*trait_effect_score_scaled
  )

  hymenachne_assigned_susceptibility <- hymenachne_assigned_susceptibility %>%

    # merge in non-threat components
    left_join(precariousness |> select(-threatened_status)) |>
    left_join(regeneration_ability |> select(-threatened_status)) |>

    # for precariousness and regeneration, assign mean value for missing data
    mutate(
      precariousness_susceptibility = ifelse(is.na(precariousness_susceptibility),
                                   mean(precariousness_susceptibility, na.rm = TRUE),
                                   precariousness_susceptibility),
      
      regeneration_ability_susceptibility = ifelse(is.na(regeneration_ability_susceptibility),
                                   mean(regeneration_ability_susceptibility, na.rm = TRUE),
                                   regeneration_ability_susceptibility)
    )

  # rescale individual susceptibility to 0-1 across all taxa before combining with
  # regeneration/precariousness, so those two terms (fixed to a narrow +/-0.1-ish
  # range everywhere) have a consistent relative influence across every threat,
  # rather than one that depends on this threat's own raw susceptibility spread
  hymenachne_IS_range <- range(hymenachne_assigned_susceptibility$total_individual_susceptibility, na.rm = TRUE)

  hymenachne_assigned_susceptibility <- hymenachne_assigned_susceptibility %>%
    mutate(
      total_individual_susceptibility_scaled = (total_individual_susceptibility - hymenachne_IS_range[1]) / diff(hymenachne_IS_range)
    ) %>%

    # sum together 3 components, normalise
    mutate(
      population_susceptibility = (1*total_individual_susceptibility_scaled) + regeneration_ability_susceptibility + precariousness_susceptibility,
      population_susceptibility = ifelse(total_individual_susceptibility == 0, 0, population_susceptibility)
    )

  # linearly rescale population_susceptibility to 0-1 across taxa
  hymenachne_susceptibility_range <- range(hymenachne_assigned_susceptibility$population_susceptibility, na.rm = TRUE)

  hymenachne_assigned_susceptibility <- hymenachne_assigned_susceptibility %>%
    mutate(
      population_susceptibility = (population_susceptibility - hymenachne_susceptibility_range[1]) / diff(hymenachne_susceptibility_range),

    # assign susceptibility bin (individual_susceptibility_bin removed - not reported to DCCEEW)
      population_susceptibility_bin = create_susceptibility_bins(population_susceptibility, bin_thresholds = hymenachne_bin_thresholds)
    ) |>

    # confidence-based plausible range around the numeric susceptibility score
    add_susceptibility_range(
      susceptibility_col = "population_susceptibility",
      confidence_col = "susceptibility_confidence",
      bin_thresholds = hymenachne_bin_thresholds
    )

  hymenachne_assigned_susceptibility_threatened <- hymenachne_assigned_susceptibility %>%
    filter(!is.na(threatened_status)) %>%
    mutate(taxonomic_susceptibility = 0) %>%
    # coding exercise (2026-08-24): flag-only column, not used in scoring - see
    # hymenachne_known_impact_species above
    mutate(known_or_suspected_impact = ifelse(taxon_name %in% hymenachne_known_impact_species, "X", NA)) %>%
    select(taxon_name, genus, family, threatened_status,
           known_or_suspected_impact,
           total_individual_susceptibility,
           population_susceptibility, population_susceptibility_bin, susceptibility_confidence,
           population_susceptibility_min, population_susceptibility_max,
           population_susceptibility_bin_min, population_susceptibility_bin_max,
           habitat, habitat_susceptibility, trait_effect_score_scaled,
           taxonomic_susceptibility, prop_traits_scored,
           contains(c("susceptibility_effect_score")), everything()) %>%
    arrange(-population_susceptibility) |>
    write_csv("~/GitHub/threatened_species_traits/export/hymenachne_susceptibility_scores.csv", na = "")
```

#### - summary table of scorings

```{r, echo=FALSE, warning=FALSE, message=FALSE}
hymenachne_trait_effects %>%
  arrange(effect_score, trait) %>% 
  kable(caption = "Trait effect scores for hymenachne.") %>% 
  kable_classic()

hymenachne_taxonomic_effects %>%
  arrange(-effect_score, taxon) %>% 
  kable(caption = "Taxa with explicit taxonomic effect scores for hymenachne.") %>% 
  kable_classic()

habitat_scores %>%
  select(habitat = trait_value, effect_score = hymenachne) %>%
  group_by(effect_score) %>%
  mutate(habitat = paste0(habitat, collapse = "; ")) %>%
  ungroup() %>%
  distinct() %>% 
  select(habitat, effect_score) %>%
  arrange(effect_score) %>%
  kable(caption = "habitat effect scores for Hymenachne") %>% 
  kable_classic()

hymenachne_assigned_susceptibility_threatened |>
  count(population_susceptibility_bin) |>
  mutate(pct = round(100 * n / sum(n), 0)) |>
  arrange(desc(population_susceptibility_bin)) |>
  left_join(bin_thresholds_table(hymenachne_bin_thresholds), by = "population_susceptibility_bin") |>
  kable(caption = "taxa in each population susceptibility bin", col.names = c("Population susceptibility bin", "n", "%", "Threshold range")) %>%
  kable_classic()
```

#### - susceptibility scores density plot

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.width=10, fig.height=13}
df_long_scaled <- hymenachne_assigned_susceptibility %>%
  select(`habitat susceptibility` = habitat_susceptibility,
         `trait and taxonomy susceptibility` = trait_effect_score_scaled,
         `combined individual susceptibility` = total_individual_susceptibility,
         `mean population susceptibility` = population_susceptibility,
         `min population susceptibility` = population_susceptibility_min,
         `max population susceptibility` = population_susceptibility_max,
         threatened_status) %>%
  pivot_longer(cols = c(1:6),
               names_to = "category", values_to = "susceptibility")

make_8_panel_susceptibility_density_plot(df_long_scaled, paste0("hymenachne (sigmoidal_rescale, k = ", sigmoidal_rescale_k, ")"))
```

#### - save 4-panel report figure to PNG

```{r, echo=FALSE, warning=FALSE, message=FALSE, results='hide', include=FALSE}
png("~/GitHub/threatened_species_traits/export/figures/hymenachne_4panel.png", width = 9, height = 6.5, units = "in", res = 150)
make_4_panel_susceptibility_density_plot(
  hymenachne_assigned_susceptibility, "hymenachne",
  panel_cols = c("habitat_susceptibility", "trait_effect_score_scaled", "total_individual_susceptibility", "population_susceptibility"),
  panel_labels = c("habitat susceptibility", "trait and taxonomy susceptibility", "individual susceptibility", "overall population susceptibility"),
  x_min = 0,
  x_max = 1.2
)
dev.off()
```

# Long results, by threat

## Phytophthora cinnamomi

```{r, echo=FALSE, message=FALSE, warning=FALSE}
phytophthora_assigned_susceptibility_threatened |>
  slice(1:50) |>
  select(family, taxon_name, phytophthora_cinnamomi_susceptibility, population_susceptibility, population_susceptibility_bin, population_susceptibility_bin_min, population_susceptibility_bin_max) |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  kable(caption = "50 taxa scored at highest susceptibility from phytophthora", col.names = c("Family", "Species", "Measured susceptibility", "Population susceptibility", "Pop'n susceptibility bin", "Pop'n susceptibility bin, min",  "Pop'n susceptibility bin, max")) %>%
  kable_classic()
```

## Myrtle rust 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
myrtle_rust_assigned_susceptibility_threatened |>
  slice(1:50) |>
  select(family, taxon_name, myrtle_rust_susceptibility, population_susceptibility, population_susceptibility_bin, population_susceptibility_bin_min, population_susceptibility_bin_max) |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  kable(caption = "50 taxa scored at highest susceptibility from myrtle rust", col.names = c("Family", "Species", "Measured susceptibility", "Population susceptibility", "Pop'n susceptibility bin", "Pop'n susceptibility bin, min",  "Pop'n susceptibility score, max")) %>%
  kable_classic()
```

## Goats

```{r, echo=FALSE, message=FALSE, warning=FALSE}
goats_assigned_susceptibility_threatened |>
  slice(1:50) |>
  select(family, taxon_name, population_susceptibility, population_susceptibility_bin, population_susceptibility_bin_min, population_susceptibility_bin_max) |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  kable(caption = "50 taxa scored at highest susceptibility from feral goats", col.names = c("Family", "Species", "Population susceptibility", "Pop'n susceptibility score", "Pop'n susceptibility score, min",  "Pop'n susceptibility score, max")) %>%
  kable_classic()
```

## Deer

```{r, echo=FALSE, message=FALSE, warning=FALSE}
deer_assigned_susceptibility_threatened |>
  slice(1:50) |>
  select(family, taxon_name, population_susceptibility, population_susceptibility_bin, population_susceptibility_bin_min, population_susceptibility_bin_max) |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  kable(caption = "50 taxa scored at highest susceptibility from feral deer", col.names = c("Family", "Species", "Population susceptibility", "Pop'n susceptibility score", "Pop'n susceptibility score, min",  "Pop'n susceptibility score, max")) %>%
  kable_classic()
```

## Pigs

```{r, echo=FALSE, message=FALSE, warning=FALSE}
pigs_assigned_susceptibility_threatened |>
  slice(1:50) |>
  select(family, taxon_name, population_susceptibility, population_susceptibility_bin, population_susceptibility_bin_min, population_susceptibility_bin_max) |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  kable(caption = "50 taxa scored at highest susceptibility from feral pigs", col.names = c("Family", "Species", "Population susceptibility", "Pop'n susceptibility score", "Pop'n susceptibility score, min",  "Pop'n susceptibility score, max")) %>%
  kable_classic()
```

## Perennial C4 grasses

```{r, echo=FALSE, message=FALSE, warning=FALSE}
tropical_C4_grasses_assigned_susceptibility_threatened |>
  slice(1:50) |>
  select(family, taxon_name, population_susceptibility, population_susceptibility_bin, population_susceptibility_bin_min, population_susceptibility_bin_max) |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  kable(caption = "50 taxa scored at highest susceptibility from invasive perennial C4 grasses", col.names = c("Family", "Species", "Population susceptibility", "Pop'n susceptibility score", "Pop'n susceptibility score, min",  "Pop'n susceptibility score, max")) %>%
  kable_classic()
```

## Olive hymenachne

```{r, echo=FALSE, message=FALSE, warning=FALSE}
hymenachne_assigned_susceptibility_threatened |>
  slice(1:50) |>
  select(family, taxon_name, population_susceptibility, population_susceptibility_bin, population_susceptibility_bin_min, population_susceptibility_bin_max) |>
  mutate(across(where(is.numeric), ~round(.x, 2))) |>
  kable(caption = "50 taxa scored at highest susceptibility from olive hymenachne", col.names = c("Family", "Species", "Population susceptibility", "Pop'n susceptibility score", "Pop'n susceptibility score, min",  "Pop'n susceptibility score, max")) %>%
  kable_classic()
```

# Model validation

(written by Claude)

Back-tests the trait+habitat composite score against independently-observed ground truth, for the three threats where such data exists: Phytophthora and myrtle rust (lab/field susceptibility trials, 0-3 ordinal scale) and goats (palatability observations, 1-4 ordinal scale). This directly answers "when we deliberately don't tell the model a species' known susceptibility, how well does its trait+habitat profile predict it anyway?" - a genuine out-of-sample-style check, not a restatement of data the model was already given.

To avoid circularity, every threat here is validated on its **pre-overwrite** score - the trait/habitat-only prediction, before the known-susceptibility overwrite (see `total_individual_susceptibility_overwrite` for Phytophthora/myrtle rust) or the known-palatability multiplier (goats) is applied. Those overwrites deliberately use the same ground-truth columns being validated against here, so scoring the post-overwrite value against them would just be checking whether a copy operation worked, not whether the trait model has any real predictive power.

As a result of these checks, the overwrite was removed for goats.

Two metrics, both threshold-free where possible:

- **Spearman's rho**: do species the model scores higher also tend to rank higher on the independently-measured scale? Robust to the two scales having different shapes (ordinal susceptibility bins vs continuous susceptibility score) - only the ordering matters.
- **AUC (Mann-Whitney)**: if the known scale is split into "susceptible" vs "not" (thresholds noted per threat below), how often does the model rank a truly-susceptible species above a truly-resistant one? 0.5 = no better than random, 1.0 = perfect separation. This is the more decision-relevant number - it's the same statistic used to validate diagnostic classifiers generally.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# Mann-Whitney U formulation of AUC - dependency-free, exact for this sample size.
compute_auc <- function(scores, is_positive) {
  keep <- !is.na(scores) & !is.na(is_positive)
  scores <- scores[keep]; is_positive <- is_positive[keep]
  pos <- scores[is_positive]
  neg <- scores[!is_positive]
  if (length(pos) == 0 || length(neg) == 0) return(NA_real_)
  ranks <- rank(c(pos, neg))
  n_pos <- length(pos); n_neg <- length(neg)
  (sum(ranks[seq_len(n_pos)]) - n_pos * (n_pos + 1) / 2) / (n_pos * n_neg)
}

# Validates one threat's pre-overwrite model score against a known ground-truth column.
# `positive_threshold`: known values >= this are treated as "truly susceptible" for the
# AUC calculation; has no effect on the Spearman correlation, which is threshold-free.
validate_against_known <- function(data, score_col, known_col, threat_label, positive_threshold) {

  d <- data %>%
    transmute(
      score = .data[[score_col]],
      known = .data[[known_col]]
    ) %>%
    filter(!is.na(score), !is.na(known))

  spearman <- suppressWarnings(cor.test(d$score, d$known, method = "spearman"))
  auc <- compute_auc(d$score, d$known >= positive_threshold)

  print(
    ggplot(d, aes(x = known, y = score)) +
      geom_jitter(width = 0.05, height = 0, alpha = 0.5) +
      geom_smooth(method = "lm", se = FALSE, color = "grey20") +
      theme_classic() +
      labs(
        x = paste0(known_col, " (known, independently observed)"),
        y = paste0(score_col, " (model, trait+habitat only)"),
        title = paste0(threat_label, ": model score vs known ground truth"),
        subtitle = sprintf("Spearman rho = %.2f (p = %.3g)  |  AUC = %.2f  |  n = %d",
                            spearman$estimate, spearman$p.value, auc, nrow(d))
      )
  )

  tibble(
    threat = threat_label,
    n = nrow(d),
    spearman_rho = round(unname(spearman$estimate), 3),
    spearman_p = signif(spearman$p.value, 3),
    auc = round(auc, 3)
  )
}
```

Human commentary:
There is a high distribution of Phytophthora scores across taxa allowing species to be distributed across both x- and y-axes. In contrast most myrtle rust scores are `3` (moderate), so inherently less ability to have a strong correlation. It is not surprising there is poor model fix for the goat palatibility data, since these are mostly genus-level scores, applied to all species in the genus. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
validation_results <- bind_rows(

  validate_against_known(
    phytophthora_assigned_susceptibility, "total_individual_susceptibility", "phytophthora_cinnamomi_susceptibility",
    "Phytophthora cinnamomi", positive_threshold = 1.5
  ),

  validate_against_known(
    myrtle_rust_assigned_susceptibility, "total_individual_susceptibility", "myrtle_rust_susceptibility",
    "Myrtle rust", positive_threshold = 1.5
  ),

  # goats' exported total_individual_susceptibility is already post-palatability-multiplier - reconstruct the
  # trait+habitat-only score the same way the multiplier itself starts from
  goats_assigned_susceptibility %>%
    mutate(pre_multiplier_susceptibility = habitat_susceptibility * trait_effect_score_scaled) %>%
    validate_against_known(
      "pre_multiplier_susceptibility", "plant_goat_palatability", "Goats",
      positive_threshold = 3  # matches the >=3 "palatable" cutoff already used by the model's own multiplier
    )
)

validation_results %>%
  kable(caption = "Model validation: trait+habitat score vs independently-known susceptibility/palatability, pre-overwrite.") %>%
  kable_classic()
```

**Reading this table**: a positive Spearman rho means the model and the independent observations agree on ranking; AUC above 0.5 means the model separates known-susceptible from known-resistant species better than chance. Neither metric being high is itself a finding worth acting on - it's telling you how much the trait/habitat data (as opposed to the taxonomic overrides and known-value overwrites layered on top) is actually carrying the model's predictive weight for the taxa where we can check.
